---
title: "Multiple Regression Analysis: Estimation"
format: html
---

```{r}
#| label: 'setup'
#| message: false
#| warning: false

library(tidyverse)
library(wooldridge)
library(knitr)
library(modelsummary)
library(broom)
```

## C1 \[BWGHT\]

> A problem of interest to health officials (and others) is to determine the effects of smoking during pregnancy on infant health. One measure of infant health is birth weight; a birth weight that is too low can put an infant at risk for contracting various illnesses. Since factors other than cigarette smoking that affect birth weight are likely to be correlated with smoking, we should take those factors into account. For example, higher income generally results in access to better prenatal care, as well as better nutrition for the mother. An equation that recognizes this is $$bwght = \beta_0 + \beta_1 cigs + \beta_2 faminc + u$$

```{r}
#| label: c1_data

c1 <- bwght
kable(head(c1))
```

> (i) What is the most likely sign for $\beta_2$?

**Interpretation:**

$\beta_2$ is expected to be **positive**. Higher family income $\rightarrow$ better nutrition $\rightarrow$ higher birth weight

> (ii) Do you think `cigs` and `faminc` are likely to be correlated? Explain why the correlation might be positive or negative.

**Interpretation:**

Yes, **negative** correlation. smoking rates is higher among lower-income demographics. Omitting family income will cause Omitted Variable Bias (OVB).

> (iii) Now, estimate the equation with and without `faminc`, using the data in BWGHT. Report the results in equation form, including the sample size and R-squared. Discuss your results, focusing on whether adding `faminc` substantially changes the estimated effect of `cigs` on `bwght`.

```{r}
#| label: c1_iii

c1m <- lm(bwght ~ cigs, data = c1)
c1m2 <- lm(bwght ~ cigs + faminc, data = c1)

c1models <- list(
  "Without Income" = c1m,
  "With Income" = c1m2
)

modelsummary(
  c1models,
  fmt = 2,
  statistic = NULL,
  gof_map = c("nobs", "r.squared"),
  coef_rename = c(
    "(Intercept)" = "Constant",
    "cigs" = "Cigarettes/Day",
    "faminc" = "Family Income ($1000s)"
  )
)
```

**Interpretation:**

Including income reduces the smoking coefficient from approximately $-0.514$ to $-0.463$. This indicates that the simple regression suffered from omitted variable bias.$$ \text{Bias} = \beta_2 \times \tilde \delta_1$$

1\. $\beta_2$: Positive, rich people have heavier babies.

2\. $\tilde \delta_1$: Negative, poor people smoke more

3\. Effect: Positive $\times$ Negative = **Negative Bias**

4\. $\beta_1$ is negative, negative bias will increase it from -0.46 to -0.51

## C2 \[HPRICE1\]

> Use the data in HPRICE1 to estimate the model $$price = \beta_0 + \beta_1sqrft + \beta_2bdrms + u$$ where `price` is the house price measured in thousands of dollars.

```{r}
#| label: c2_data

c2 <- hprice1
kable(head(c2))
```

> (i) Write out the results in equation form.

```{r}
#| label: c2_i

c2m <- lm(price ~ sqrft + bdrms, data = c2)

modelsummary(
  c2m,
  fmt = 2,
  statistic = NULL,
  gof_map = c("nobs", "r.squared"),
  coef_rename = c(
    "(Intercept)" = "Constant",
    "sqrft"       = "Size (sq. ft.)",
    "bdrms"       = "Number of Bedrooms"
  )
)
```

> (ii) What is the estimated increase in price for a house with one more bedroom, holding square footage constant?

**Interpretation:**

Holding square footage constant, adding one more bedroom increases price by \$15,200. This estimate is not useful because adding small rooms without adding square footage will make it feel cramped.

> (iii) What is the estimated increase in price for a house with an additional bedroom that is 140 square feet in size? Compare this to your answer in part (ii).

```{r}
#| label: c2_iii

base_house <- data.frame(sqrft = 140, bdrms = 1)
upgraded_house <- data.frame(sqrft = 280, bdrms = 2)

price_1 <- predict(c2m, newdata = base_house)
price_2 <- predict(c2m, newdata = upgraded_house)

cat(round(price_2 - price_1, 2))
```

**Interpretation:**

The estimates increase is **\$33,180**. This represents the combined value of adding a room that is 140 square feet $$
140 \times 0.13 + 15.2 = 33.2$$

> (iv) What percentage of the variation in price is explained by square footage and number of bedrooms?

**Interpretation:**

**63.2%** of the variation in house prices are explained by the two variables.

> (v) The first house in the sample has `sqrft` = 2,438 and `bdrms` = 4. Find the predicted selling price

```{r}
#| label: c2_v

cat(c2m$fitted.values[1])
```

**Interpretation:**

Predicted selling price is **\$354,605**

> (vi) The actual selling price of the first house in the sample was \$300,000 Find the residual for this house. Does it suggest that the buyer underpaid or overpaid for the house?

```{r}
#| label: c2_vi

cat(c2m$residuals[1])
```

**Interpretation:**

Residual is **\$-54,605**. The buyer underpaid for the house.

## C3 \[CEOSAL2\]

> The file CEOSAL2 contains data on 177 chief executive officers and can be used to examine the effects of firm performance on CEO salary.

```{r}
#| label: c3_data

c3 <- ceosal2
kable(head(c3))
```

> (i) Estimate a model relating annual salary to firm sales and market value. Make the model of the constant elasticity variety for both independent variables.

```{r}
#| label: c3_i

c3m <- lm(lsalary ~ lsales + lmktval, data = c3)

modelsummary(
  c3m,
  fmt = 2,
  statistic = NULL,
  gof_map = c("nobs", "r.squared"),
  coef_rename = c(
    "(Intercept)" = "Constant",
    "lsales" = "Log(Sales)",
    "lmktval" = "Log(Market Value)"
  )
)
```

> (ii) Add `profits` to the model from part (i). Why can this variable not be included in logarithmic form? Would you say that these firm performance variables explain most of the variation in CEO salaries?

```{r}
#| label: c3_ii

ggplot(c3, aes(x = profits)) +
  geom_histogram(bins = 30)
```

```{r}
#| label: c3_ii2

c3m2 <- lm(lsalary ~ lsales + lmktval + profits, data = c3)

modelsummary(
  c3m2,
  fmt = 2,
  statistic = NULL,
  gof_map = c("nobs", "r.squared"),
  coef_rename = c(
    "(Intercept)" = "Constant",
    "lsales" = "Log(Sales)",
    "lmktval" = "Log(Market Value)",
    "profits" = "Profits"
  )
)
```

**Interpretation:**

Profits can't be added in log form because it has non positive values. The model explains **30%** of the variation in the salaries. The remaining 70% is determined by other factors like tenure, unobserved CEO talent, or corporate politics.

> (iii) Add the variable `ceoten` to the model in part (ii). What is the estimated percentage return for another year of CEO tenure, holding other factors fixed?

```{r}
#| label: c3_iii

c3m3 <- lm(
  lsalary ~ lsales + lmktval + profits + ceoten,
  data = c3
)

modelsummary(
  c3m3,
  fmt = 2,
  statistic = NULL,
  gof_map = c("nobs", "r.squared"),
  coef_rename = c(
    "(Intercept)" = "Constant",
    "lsales" = "Log(Sales)",
    "lmktval" = "Log(Market Value)",
    "profits" = "Profits",
    "ceoten" = "Tenure"
  )
)
```

**Interpretation:**

$\beta_4$ = 0.01, one additional year of tenure is associated with **1%** increase in salary, holding firm performance constant.

> (iv) Find the sample correlation coefficient between the variables `log(mktval)` and `profits`. Are these variables highly correlated? What does this say about the OLS estimators?

```{r}
#| label: c3_iv

c3 |>
  summarise(
    Correlation = cor(lmktval, profits)
  ) |>
  round(2)
```

**Interpretation:**

Correlation = **0.78**. This high correlation between market value and profits indicates multicollinearity. While this does not bias our estimates or inflate the overall $R^2$, it makes it difficult to isolate the individual partial effects of profits versus market value.

## C4 \[ATTEND\]

> Use the data in ATTEND for this exercise.

```{r}
#| label: c4_data

c4 <- attend
kable(head(c4))
```

> (i) Obtain the minimum, maximum, and average values for the variables `atndrte`, `priGPA`, and `ACT`.

```{r}
#| label: c4_i

c4 |>
  summarise(
    across(
      c(atndrte, priGPA, ACT),
      list(
        min = \(x) min(x, na.rm = TRUE),
        max = \(x) max(x, na.rm = TRUE),
        avg = \(x) mean(x, na.rm = TRUE)
      )
    )
  ) |>
  round(2)
```

**Interpretation:**

Attendance rate lies in the range \[6.25, 100\] with an average of 81.71. Prior GPA is in the interval \[0.86, 3.93\] with an average of 2.59. ACT score is in the interval \[13, 32\] with an average of 22.51.

> (ii) Estimate the model $$atndrte = \beta_0 + \beta_1 priGPA + \beta_2 ACT + u$$ and write the results in equation form. Interpret the intercept. Does it have a useful meaning?

```{r}
#| label: c4_ii

c4m <- lm(atndrte ~ priGPA + ACT, data = c4)

modelsummary(
  c4m,
  fmt = 2,
  statistic = NULL,
  gof_map = c("nobs", "r.squared"),
  coef_map = c(
    "(Intercept)" = "Constant",
    "priGPA"      = "Prior GPA",
    "ACT"         = "ACT Score"
  )
)
```

```{r}
#| label: c4_ii_2

c4 |>
  select(atndrte, priGPA, ACT) |>
  pivot_longer(cols = c(priGPA, ACT), names_to = "variable", values_to = "value") |>
  ggplot(aes(x = value, fill = atndrte > 80)) +
  geom_histogram(bins = 20, position = "identity", alpha = 0.6) +
  facet_wrap(~variable, scales = "free") +
  theme_minimal() +
  scale_x_continuous(n.breaks = 10) +
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust = 1))
```

**Interpretation:**

$\hat \beta_0 = 75.7$. A student with zero ACT score and zero prior GPA is expected to have attendance rate = **75.7**. The intercept has no useful meaning because regression is extrapolating.

> (iii) Discuss the estimated slope coefficients. Are there any surprises?

**Interpretation**

ACT score has a slope of **-1.72**. This is surprising as a higher score is usually correlated with higher attendance.

> (iv) What is the predicted atndrte if `priGPA` = 3.65. and `ACT` = 20? What do you make of this result? Are there any students in the sample with these values of the explanatory variables?

```{r}
#| label: c4_iv

c4 |>
  filter(ACT == 20 & priGPA == 3.65) |>
  summarise(
    Count = n()
  )
```

```{r}
#| label: c4_iv_2

df <- data.frame(priGPA = 3.65, ACT = 20)

prediction <- predict(c4m, newdata = df)
cat(prediction)
```

**Interpretation:**

A student with prior GPA of 3.65 and ACT score of 20 is predicted to have attendance rate of **104**; which passes maximum value of 100. The model has no students with such values so it is extrapolating.

> (v) If Student A has `priGPA` = 3 1. and `ACT` = 21 and Student B has `priGPA = 2.1`. and `ACT = 26`, what is the predicted difference in their attendance rates?

```{r}
#| label: c4_v

dfA <- data.frame(priGPA = 3.1, ACT = 21)
dfB <- data.frame(priGPA = 2.1, ACT = 26)

predictionA <- predict(c4m, newdata = dfA)
predictionB <- predict(c4m, newdata = dfB)

cat(predictionA - predictionB)
```

**Interpretation:**

The expected difference in attendance rate between the two students is **25.8**.

## C5 \[WAGE1\]

> (i) Confirm the partialling out interpretation of the OLS estimates by explicitly doing the partialling out for Example 3.2. $$\widehat{\log(wage)} = 0.284 + 0.092 educ + 0.0041 exper + 0.022 tenure$$ This first requires regressing `educ` on `exper` and `tenure` and saving the residuals, $\hat u$ Then, regress `log(wage)` on $\hat u$. Compare the coefficient on $\hat u$ with the coefficient on educ in the regression of log(wage) on educ, exper, and tenure.

```{r}
#| label: c5_data

c5 <- wage1
kable(head(c5))
```

```{r}
#| label: c5_i

c5m <- lm(educ ~ exper + tenure, data = c5)

c5 <- c5 |>
  mutate(
    u_hat = resid(c5m)
  )

c5m2 <- lm(lwage ~ u_hat, data = c5)

modelsummary(
  c5m2,
  fmt = 2,
  statistic = NULL,
  gof_map = c("nobs", "R.square"),
  coef_map = c(
    "(Intercept)" = "Constant",
    "u_hat" = "Residuals"
  )
)
```

**Interpretation:**

Slope of the residual term is same as that of education **(0.09)**. This proves that multiple regression computes partial effect of each independent variable.

::: callout-note
### Frisch-Waugh-Lovell Theorem

In a multiple regression, the coefficient for $x_1$ represents partial effect of $x_1$ on $y$ after removing overlop with other variables:

1.  Regress: $x_1 = \alpha_0 + \alpha_1 x_2 + \alpha_2 x_3 + u$
2.  Extract: $\hat{u}$ (This is "$x_1$" that is completely uncorrelated with $x_2$ and $x_3$).
3.  Result: The slope $\beta_1$ from $y = \beta_0 + \beta_1 \hat{u} + e$ is identical to the coefficient in the full multiple regression.
4.  The Formula$$\hat{\beta}_{x_1} = \frac{\sum \hat{u}_i y_i}{\sum \hat{u}_i^2}$$
:::

## C6 [WAGE2]
