---
title: "The Simple Regression Model"
format: html
---

```{r}
#| label: 'setup'
#| message: false
#| warning: false

library(tidyverse)
library(wooldridge)
library(knitr)
library(modelsummary)
library(broom)
```

## C1 \[401K\]

> The data in 401K are a subset of data analyzed by Papke (1995) to study the relationship between participation in a 401(k) pension plan and the generosity of the plan. The variable `prate` is the percentage of eligible workers with an active account; this is the variable we would like to explain. The measure of generosity is the plan match rate, `mrate`. This variable gives the average amount the firm contributes to each worker’s plan for each \$1 contribution by the worker. For example, if `mrate` = 0 50. , then a \$1 contribution by the worker is matched by a 50¢ contribution by the firm.

```{r}
#| label: c1_data

c1 <- k401k
kable(head(c1))
```

> (i) Find the average participation rate and the average match rate in the sample of plans.

```{r}
#| label: c1_i

c1 |>
  summarise(
    Avg_Part_Rt = mean(prate),
    Avg_Match_Rt = mean(mrate)
  ) |>
  round(2)
```

**Interpretation:**

On average, **87.36%** of eligible employees participate in their company's 401(k) plan. Their company matches **$0.73** for every dollar an employee contributes.

> estimate the simple regression equation $$\widehat{prate} = \hat \beta_0 + \hat \beta_1 mrate$$ and report the results along with the sample size and R-squared.

```{r}
#| label: c1_ii

c1m <- lm(prate ~ mrate, data = c1)

modelsummary(
  c1m,
  fmt = 2,
  statistic = NULL,
  gof_map = c("nobs", "r.squared"),
  coef_rename = c("mrate" = "Match Rate ($)", "(Intercept)" = "Constant")
)
```

**Interpretation:**

$$
\begin{align}
\hat \beta_0 &= 83.03 & \hat\beta_1 &= 5.86 \\
R^2 &= 0.075 & n &= 1534
\end{align}
$$

> (iii) Interpret the intercept in your equation. Interpret the coefficient on `mrate`.

**Interpretation:**

If the match rate is 0, the predicted participation rate is **83.08%**. For every $1 increase in the match rate, the participation rate is predicted to increase by **5.86** percentage points

> (iv) Find the predicted prate when `mrate` = 3.5. Is this a reasonable prediction? Explain what is happening here.

```{r}
#| label: c1_iv

new_data <- data.frame(mrate = 3.5)

c1m |>
  predict(newdata = new_data) |>
  round(2)
```

::: callout-note
### Level-Level model prediction

To predict $\hat y$ value given a new value of x in a level-level model, substitute in this formula

$$\hat y = \hat \beta_0 + \hat \beta_1 x$$
:::

**Interpretation:**

The model predicts a participation rate of **103.59%**, which exceeds the maximum logical value of 100%. This highlights the danger of extrapolation and the limitation of simple linear models when dealing with bounded dependent variables (like percentages or probabilities).

> (v) How much of the variation in `prate` is explained by `mrate`? Is this a lot in your opinion?

**Interpretation:**

The match rate explains only about 7.5% of the variation in the participation rate. This is relatively low, meaning the remaining 92.5% of the variation is due to other unobserved factors (like employee age, income, or financial literacy).

## C2 \[CEOSAL2\]

> The data set in CEOSAL2 contains information on chief executive officers for U.S. corporations. The variable `salary` is annual compensation, in thousands of dollars, and `ceoten` is prior number of years as company CEO.

```{r}
#| label: c2_data

c2 <- ceosal2
kable(head(c2))
```

> (i) Find the average salary and the average tenure in the sample.

```{r}
#| label: c2_i

c2 |>
  summarise(
    Avg_Salary = mean(salary),
    Avg_Tenure = mean(ceoten)
  ) |>
  round(2)
```

**Interpretation:**

The average annual compensation for a CEO in this sample is $865,860, with an average tenure in the CEO position of 7.95 years

> (ii) How many CEOs are in their first year as CEO (that is, `ceoten` = 0)? What is the longest tenure as a CEO?

```{r}
#| label: c2_ii

c2 |>
  summarise(
    Fresh_CEO = sum(ceoten == 0),
    Veteran = max(ceoten)
  )
```

**Interpretation:**

There are 5 CEOs currently in their first year on the job (ceoten = 0), while the longest-serving veteran in the sample has been CEO for 37 years

> (iii) Estimate the regression model $$ \log(salary) = \beta_0 + \beta_1 ceoten + u$$ and report your results in the usual form. What is the (approximate) predicted percentage increase in salary given one more year as a CEO?

```{r}
#| label: c2_iii

c2m <- lm(lsalary ~ ceoten, data = c2)

modelsummary(
  c2m,
  fmt = 2,
  statistic = NULL,
  gof_map = c("nobs", "r.squared"),
  coef_rename = c("ceoten" = "Tenure", "(Intercept)" = "Constant")
)
```

**Interpretation:**

Because this is a log-level model, we multiply the slope coefficient by 100 to get the approximate percentage change. For every 1 additional year of tenure as a CEO, annual salary is predicted to increase by approximately 1%.

## C3 \[SLEEP75\]

> Use the data in SLEEP75 from Biddle and Hamermesh (1990) to study whether there is a tradeoff between the time spent sleeping per week and the time spent in paid work. We could use either variable as the dependent variable. For concreteness, estimate the model $$sleep = \beta_0 + \beta_1 totwrk + u$$ where `sleep` is minutes spent sleeping at night per week and `totwrk` is total minutes worked during the week.

```{r}
#| label: c3_data

c3 <- sleep75
kable(head(c3))
```

> (i) Report your results in equation form along with the number of observations and $R^2$. What does the intercept in this equation mean?

```{r}
#| label: c3_i

c3m <- lm(sleep ~ totwrk, data = c3)

modelsummary(
  c3m,
  fmt = 2,
  statistic = NULL,
  gof_map = c("nobs", "r.squared"),
  coef_rename = c(
    "(Intercept)" = "Constant",
    "totwrk" = "Minutes Worked per Week"
  )
)
```

**Interpretation:**

The intercept indicates that for a person who works zero minutes during the week, their predicted sleep is 3586 minutes per week (approximately 8.54 hours per night). The slope coefficient indicates that for every additional 100 minutes worked per week, weekly sleep is predicted to decrease by 15 minutes.

> (ii) If `totwrk` increases by 2 hours, by how much is `sleep` estimated to fall? Do you find this to be a large effect?

```{r}
#| label: c3_ii

sleep_fall <- coef(c3m)["totwrk"] * 120

cat("Estimated change in sleep:", round(sleep_fall, 2), "minutes")
```

| Model | Regression Equation | Interpretation of $\hat \beta_1$ | Name |
|------------------|------------------|-------------------|------------------|
| Level-Level | $y = \hat \beta_0 + \hat \beta_1 x$ | $\Delta \hat y = \hat \beta \Delta x$ | Marginal Effect |
| Log-Level | $\log(y) = \hat \beta_0 + \hat \beta_1 x$ | $\% \Delta \hat y \approx (100\hat \beta_1) \Delta x$ | Semi-elasticity |
| Level-Log | $y = \hat \beta_0 + \hat \beta_1 \log(x)$ | $\Delta \hat y \approx (\hat \beta_1)/100 \% \Delta x$ | \- |
| Log-Log | $\log(y) = \hat \beta_0 + \hat \beta_1 \log(x)$ | $\% \Delta \hat y \approx \hat \beta_1 \% \Delta x$ | Elasticity |

**Interpretation:**

An increase of two hours (120 minutes) of work per week is associated with a predicted reduction in weekly sleep of 18 minutes (or about 2.6 minutes per night). While this estimate might be statistically significant, it is practically (economically) insignificant. Losing less than 3 minutes of sleep per night has no meaningful impact on a person's life or productivity.

## C4 \[WAGE2\]

> Use the data in WAGE2 to estimate a simple regression explaining monthly salary (`wage`) in terms of IQ score (`IQ`).

```{r}
#| label: c4_data

c4 <- wage2
kable(head(c4))
```

> (i) Find the average salary and average IQ in the sample. What is the sample standard deviation of IQ? (IQ scores are standardized so that the average in the population is 100 with a standard deviation equal to 15).

```{r}
#| label: c4_i

c4 |>
  summarise(
    Avg_Salary = mean(wage),
    Avg_IQ = mean(IQ),
    SD_IQ = sd(IQ)
  ) |>
  round(2)
```

**Interpretation:**

The sample average monthly wage is $957.95, with an average IQ of 101.28. The sample standard deviation for IQ is 15.05

> (ii) Estimate a simple regression model where a one-point increase in `IQ` changes `wage` by a constant dollar amount. Use this model to find the predicted increase in `wage` for an increase in `IQ` of 15 points. Does `IQ` explain most of the variation in `wage`?

```{r}
#| label: c4_ii

c4m <- lm(wage ~ IQ, data = c4)

modelsummary(
  c4m,
  fmt = 2,
  statistic = NULL,
  gof_map = c("nobs", "r.squared"),
  coef_rename = c(
    "(Intercept)" = "Constant"
  )
)
```

```{r}
#| label: c4_ii_2

IQ_inc <- coef(c4m)["IQ"] * 15

cat("Estimated change in wage:", round(IQ_inc, 2), "dollars")
```

**Interpretation:**

For every 1-point increase in IQ, the predicted monthly wage increases by approximately \$8.30. Therefore, a 15-point increase in IQ is associated with a predicted wage increase of $124.55.
IQ explains only 9.6% of the variation in wage ($R^2 = 0.096$); the remaining 90.4% of the variation is unexplained by this simple model and is due to other factors.

> (iii) Now, estimate a model where each one-point increase in `IQ` has the same percentage effect on `wage`. If `IQ` increases by 15 points, what is the approximate percentage increase in predicted `wage`?

```{r}
#| label: c4_iii

c4m2 <- lm(lwage ~ IQ, data = c4)

modelsummary(
  c4m2,
  fmt = 3,
  statistic = NULL,
  gof_map = c("nobs", "r.squared"),
  coef_rename = c(
    "(Intercept)" = "Constant"
  )
)
```

```{r}
#| label: c4_iii_2

IQ_inc <- coef(c4m2)["IQ"] * 15 * 100

cat("Estimated change in wage:", round(IQ_inc, 2), "percent")
```

**Interpretation:**

Because this is a log-level model, we multiply the slope by 100 to get the percentage effect. Each 1-point increase in IQ is associated with a predicted wage increase of 0.9%. Consequently, a 15-point increase in IQ will yield a predicted wage increase of approximately 13.5%

## C5 \[RDCHEM\]

> For the population of firms in the chemical industry, let `rd` denote annual expenditures on research and development, and let `sales` denote annual sales (both are in millions of dollars).

```{r}
#| label: c5_data

c5 <- rdchem
kable(head(c5))
```

> (i) Write down a model (not an estimated equation) that implies a constant elasticity between `rd` and `sales`. Which parameter is the elasticity?

**Interpretation:**

The population model is:$$\log(rd) = \beta_0 + \beta_1 \log(sales) + u$$The parameter $\beta_1$ represents the constant elasticity of R&D expenditures with respect to firm sales.

> (ii) Now, estimate the model using the data in RDCHEM. Write out the estimated equation in the usual form. What is the estimated elasticity of `rd` with respect to `sales`? Explain in words what this elasticity means.

```{r}
#| label: c5_i

c5m <- lm(lrd ~ lsales, data = c5)

modelsummary(
  c5m,
  fmt = 2,
  statistic = NULL,
  gof_map = c("nobs", "r.squared"),
  coef_rename = c(
    "(Intercept)" = "Constant",
    "lsales" = "Log Sales"
  )
)
```

**Interpretation:**

The estimated elasticity of R&D with respect to sales is 1.08. This means that a 1% increase in a firm's sales is associated with a predicted 1.08% increase in its R&D expenditures. Because the elasticity is greater than 1, R&D spending increases at a slightly faster rate than sales.

## C6 \[MEAP93\]

> We used the data in MEAP93 for Example 2.12. In this exercise, we want to explore the relationship between the math pass rate (`math10`) and spending per student (`expend`).

```{r}
#| label: c6_data

c6 <- meap93
kable(head(meap93))
```

> (i) Do you think each additional dollar spent has the same effect on the pass rate, or does a diminishing effect seem more appropriate? Explain.

**Interpretation:**

A diminishing effect is more appropriate. The first few dollars spent (on basic necessities like textbooks or qualified teachers) likely have a large impact on the pass rate. However, once a school is well-funded, spending additional money yields smaller marginal improvements. Furthermore, because the pass rate is strictly bounded at 100%, a constant linear effect would eventually and unrealistically predict pass rates exceeding 100% for very high-spending schools.

> (ii) In the population model $$math10 = \beta_0 + \beta_1 \log(expend) + u$$ argue that $β_1 / 10$ is approximately the percentage point change in `math10` given a 10% increase in `expend`.

**Interpretation:**

$$
\begin{align}
\Delta math10 &= \beta_1 \cdot \Delta \log(expend) \\
\Delta math10 &\approx \beta_1 \cdot \left( \frac{\Delta expend}{expend} \right) \\
\Delta math10 &\approx \beta_1 \cdot (0.10) \\
\Delta math10 &\approx \frac{\beta_1}{10}
\end{align}
$$

Therefore, a 10% increase in spending ($\frac{\Delta expend}{expend} = 0.10$) results in an approximate change in the math pass rate of $\frac{\beta_1}{10}$.

> (iii) Use the data in MEAP93 to estimate the model from part (ii). Report the estimated equation in the usual way, including the sample size and R-squared.

```{r}
#| label: c6_iii

c6m <- lm(math10 ~ lexpend, data = c6)

modelsummary(
  c6m,
  fmt = 2,
  statistic = NULL,
  gof_map = c("nobs", "r.squared"),
  coef_rename = c(
    "(Intercept)" = "Constant",
    "lexpend" = "Student Spending"
  )
)
```

> (iv) How big is the estimated spending effect? Namely, if spending increases by 10%, what is the estimated percentage point increase in `math10`?

**Interpretation:**

The estimated coefficient on $\log(expend)$ is $\hat{\beta}_1 \approx 11.16$.
If spending increases by 10%, the math pass rate is predicted to increase by approximately 1.116 percentage points ($11.16 / 10$).

> (v) One might worry that regression analysis can produce fitted values for `math10` that are greater than 100. Why is this not much of a worry in this data set?

```{r}
#| label: c6_v

c6m |>
  augment() |>
  select(math10, .fitted) |>
  pivot_longer(everything()) |>
  ggplot(aes(x = value)) +
  geom_histogram(bins = 40) +
  facet_wrap(~name)
```

**Interpretation:**

We do not have to worry about the model predicting impossible values (greater than 100) for this specific sample. As shown in the histogram, the highest predicted pass rate ($\hat{y}$) generated by our model is around 30%, well below the theoretical ceiling of 100%. The maximum actual pass rate in the sample is also well below 100.

## C7 \[CHARITY\]

> Use the data in CHARITY \[obtained from Franses and Paap (2001)\] to answer the following questions:

```{r}
#| label: c7_data

c7 <- charity
kable(head(c7))
```

> (i) What is the average gift in the sample of 4,268 people (in Dutch guilders)? What percentage of people gave no gift?

```{r}
#| label: c7_i

c7 |>
  summarise(
    Avg_Gift = mean(gift),
    No_Gift_Pct = mean(gift == 0) * 100
  ) |>
  round(2)
```

**Interpretation:**

The average gift in the sample of 4,268 people is 7.44 Dutch guilders. Exactly 60% of the people in the sample gave no gift (a contribution of 0).

> (ii) What is the average mailings per year? What are the minimum and maximum values?

```{r}
#| label: c7_ii

c7 |>
  summarise(
    Min_Mailing = min(mailsyear),
    Avg_Mailing = mean(mailsyear),
    Max_Mailing = max(mailsyear)
  ) |>
  round(2)
```

**Interpretation:**

The minimum number of mailings per year is 0.25 (which equates to one mailing every four years). The average is 2.05 mailings per year, and the maximum is 3.05 mailings per year.

> (iii) Estimate the model $$gift = \beta_0 + \beta_1 mailsyear + u$$ by OLS and report the results in the usual way, including the sample size and R-squared.

```{r}
#| label: c7_iii

c7m <- lm(gift ~ mailsyear, data = c7)

modelsummary(
  c7m,
  fmt = 2,
  statistic = NULL,
  gof_map = c("nobs", "r.squared"),
  coef_rename = c(
    "(Intercept)" = "Constant",
    "mailsyear" = "Mailings per Year"
  )
)
```

> (iv) Interpret the slope coefficient. If each mailing costs one guilder, is the charity expected to make a net gain on each mailing? Does this mean the charity makes a net gain on every mailing? Explain.

**Interpretation:**

For each additional mailing sent per year, the charitable gift is predicted to increase by 2.65 guilders. If each mailing costs 1 guilder, the expected net gain per additional mailing is 1.65 guilders (2.65 - 1.00). It is crucial to note that this is an average predicted net gain across the population ; it absolutely does not guarantee that the charity makes a profit on every single piece of mail sent, especially since 60% of the sample contributes nothing.

> (v) What is the smallest predicted charitable contribution in the sample? Using this simple regression analysis, can you ever predict zero for `gift`?

```{r}
#| label: c7_v


c7m |>
  augment() |>
  summarise(
    Min_Prediction = min(.fitted)
  ) |>
  round(2)
```

**Interpretation:**

The smallest predicted charitable contribution for an individual in this specific sample is 2.68 guilders, which comes from the observation with the minimum 0.25 mailings per year ($2.01 + 2.65 \times 0.25 = 2.68$).
Using this estimated equation, it is impossible to predict a zero gift. Because the number of mailings cannot be negative ($x \ge 0$), and both $\hat{\beta}_0$ and $\hat{\beta}_1$ are positive, the absolute lowest prediction this model can logically generate is the intercept itself: $\hat{\beta}_0 = 2.01$ guilders.

## C8 \[Simulation\]

> (i) Start by generating 500 observations on $x_i$—the explanatory variable—from the uniform distribution with range \[0,10\]. (Most statistical packages have a command for the Uniform(0,1) distribution; just multiply those observations by 10.) What are the sample mean and sample standard deviation of the $x_i$?

```{r}
#| label: c8_i

set.seed(42)

x_i <- runif(500, min = 0, max = 10)

cat(round(mean(x_i), 2), round(sd(x_i), 2))
```

**Interpretation:**

The sample mean of $x_i$ is 4.9 and the sample standard deviation is 2.89. These values are extremely close to the theoretical population mean ($E[x] = 5$) and population standard deviation ($\sigma_x = \sqrt{100/12} \approx 2.887$) of a Uniform(0,10) distribution, differing only slightly due to finite sampling variance.

> (ii) Randomly generate 500 errors, $u_i$, from the Normal(0,36) distribution. (If you generate a Normal(0,1), as is commonly available, simply multiply the outcomes by six.) Is the sample average of the $u_i$ exactly zero? Why or why not? What is the sample standard deviation of the $u_i$?

```{r}
#| label: c8_ii

set.seed(42)

u_i <- rnorm(500, mean = 0, sd = 6)

cat(round(mean(u_i), 2), round(sd(u_i), 2))
```

**Interpretation:**

The sample average of $u_i$ is -0.18, which is not exactly zero. While the population expectation is exactly zero ($E[u] = 0$), a finite sample of 500 random draws will inherently contain sampling error, preventing the sample mean from perfectly matching the population mean. The sample standard deviation is 5.83, which closely approximates the true population standard deviation of 6.

> (iii) Now generate $y_i$ as $$y_i = 1+2x_i + u_i \equiv \beta_0 + \beta_1 + u_i$$ that is, the population intercept is one and the population slope is two. Use the data to run the regression of $y_i$ on $x_i$. What are your estimates of the intercept and slope? Are they equal to the population values in the above equation? Explain.

```{r}
#| label: c8_iii

y_i <- 1 + (2 * x_i) + u_i

c8m <- lm(y_i ~ x_i)

modelsummary(
  c8m,
  statistic = NULL,
  gof_map = list(),
  coef_rename = c(
    "(Intercept)" = "Constant",
    "x_i" = "x"
  )
)
```

**Interpretation:**

$\hat{\beta}_0$ is estimated to be 0.901 and $\hat{\beta}_1$ is estimated to be 1.983. They differ from the true population parameters ($\beta_0 = 1, \beta_1 = 2$) due to sampling error. The intercept deviates because the sample mean of the errors ($\bar{u}$) is not exactly zero. More importantly, the slope deviates because the sample covariance between $x_i$ and $u_i$ in this specific draw is not exactly zero.

> (iv) Obtain the OLS residuals, $u_i$, and verify that equation (2.60) holds (subject to rounding error).

```{r}
#| label: c8_iv

c8m |>
  augment() |>
  summarise(
    Sum_Resid = sum(.resid),
    Sum_x_resid = sum(x_i * .resid)
  ) |>
  round(2)
```

**Interpretation:**

The sum of the residuals ($\sum \hat{u}_i$) and the sum of the product of the regressors and residuals ($\sum x_i \hat{u}_i$) are both effectively zero. This confirms that the algebraic properties of OLS hold perfectly in the sample.

> (v) Compute the same quantities in equation (2.60) but use the errors $u_i$ in place of the residuals. Now what do you conclude?

```{r}
#| label: c8_v

c8m |>
  augment() |>
  summarise(
    Sum_Error = sum(u_i),
    Sum_x_Error = sum(x_i * u_i)
  ) |>
  round(2)
```

**Interpretation:**

When using the unobserved population errors ($u_i$) instead of the OLS residuals ($\hat{u}_i$), the sums are far from zero ($\sum u_i = -90.14$ and $\sum x_i u_i = -512.34$). This highlights the fundamental mechanical difference between errors and residuals: OLS residuals are mathematically forced by the regression algorithm to sum to zero and be uncorrelated with the regressors ($x_i$), whereas population errors are unobserved random variables that will not perfectly satisfy these orthogonality conditions in a finite sample.

> (vi) Repeat parts (i), (ii), and (iii) with a new sample of data, starting with generating the $x_i$.Now what do you obtain for $\hat \beta_0, \hat \beta_1$? Why are these different from what you obtained in part (iii)?

```{r}
#| label: c8_vi

set.seed(43)

x_i <- runif(500, min = 0, max = 10)
u_i <- rnorm(500, mean = 0, sd = 6)
y_i <- 1 + (2 * x_i) + u_i

c8m2 <- lm(y_i ~ x_i)

modelsummary(
  c8m2,
  statistic = NULL,
  gof_map = list(),
  coef_rename = c(
    "(Intercept)" = "Constant",
    "x_i" = "x"
  )
)
```

**Interpretation:**

With a new random sample, the estimates change to $\hat{\beta}_0 = 1.022$ and $\hat{\beta}_1 = 2.040$. These differ from the estimates in part (iii) because every new sample draws a unique set of $x_i$ and $u_i$ values. Consequently, the new sample mean of $u_i$ and the new sample covariance between $x_i$ and $u_i$ have changed. The OLS estimator processes this new random sample variation, resulting in different deviations from the true population parameters.

::: callout-caution
### When estimate equals parameter

Formula for the slope $\hat \beta_1$: $$\hat \beta_1 = \beta_1 + \dfrac{Cov(x,u)}{Var(x)}$$ To be equal to the parameter, covariance must be = zero

Formula for $\hat \beta_0$:

$$\hat \beta_0 = \beta_0 + (\beta_1 - \hat \beta_1) \bar x + \bar u$$ To be equal to the parameter, average of error term must be exactly zero, and $\hat \beta_1 = \beta_1$
:::

## C9 \[COUNTYMURDERS\]

> Use the data in COUNTYMURDERS to answer these questions. Use only the data for 1996.

```{r}
c9 <- countymurders |>
  filter(year == 1996)

kable(head(c9))
```

> (i) How many counties had zero murders in 1996? How many counties had at least one execution? What is the largest number of executions?

```{r}
#| label: c9_i

c9 |>
  summarise(
    Zero_murders = sum(murders == 0),
    Execution = sum(execs > 0)
  )
```

**Interpretation:**

In 1996, 1,051 counties had zero murders. There were 31 counties that had at least one execution. The largest number of executions in a single county was 3.

> (ii) Estimate the equation $$murders = \beta_0 + \beta_1 execs + u$$ by OLS and report the results in the usual way, including sample size and R-squared.

```{r}
#| label: c9_ii

c9m <- lm(murders ~ execs, data = c9)

modelsummary(
  c9m,
  fmt = 2,
  statistic = NULL,
  gof_map = c("nobs", "r.squared"),
  coef_rename = c(
    "(Intercept)" = "Constant",
    "execs" = "Execution"
  )
)
```

> (iii) Interpret the slope coefficient reported in part (ii). Does the estimated equation suggest a deterrent effect of capital punishment?

For every one additional execution in a county, the model predicts an increase of approximately 59 murders. This estimated equation does not suggest a deterrent effect of capital punishment. If executions had a deterrent effect, we would expect the slope coefficient to be negative (indicating that more executions are associated with fewer murders).

> (iv) What is the smallest number of murders that can be predicted by the equation? What is the residual for a county with zero executions and zero murders?

```{r}
#| label: c9_iv

df <- data.frame(execs = 0)

min_pred <- round(predict(c9m, newdata = df), 2)
residual <- round(0 - min_pred, 2)

cat(min_pred, residual)
```

**Interpretation:**

The minimum possible number of predicted murders from this model is the intercept, 5.46, which occurs when a county has zero executions. For a county with exactly zero executions and zero murders, the residual is $\hat{u} = y - \hat{y} = 0 - 5.46 = \mathbf{-5.46}$.

> (v) Explain why a simple regression analysis is not well suited for determining whether capital punishment has a deterrent effect on murders.

**Interpretation:**

A simple regression analysis is entirely unsuited for capturing the causal deterrent effect of capital punishment because it heavily violates the zero conditional mean assumption ($E[u|x] = 0$).
First, there is massive omitted variable bias. The error term contains unobserved factors like local poverty rates, police funding, and demographics that are correlated with both murder rates and the likelihood of executions.
Second, there is a clear issue of simultaneous (reverse) causality. Executions do not randomly occur; counties with historically high murder rates are mechanically more likely to sentence criminals to death. Therefore, murders cause executions just as much as (or more than) executions might deter murders, resulting in a positive, heavily biased slope coefficient.

## C10 \[CATHOLIC\]

> The data set in CATHOLIC includes test score information on over 7,000 students in the United States who were in eighth grade in 1988. The variables `math12` and `read12` are scores on twelfth grade standardized math and reading tests, respectively.

```{r}
#| label: c10_data

c10 <- catholic
kable(head(c10))
```

> (i) How many students are in the sample? Find the means and standard deviations of `math12` and `read12`.

```{r}
#| label: c10_i

c10 |>
  summarise(
    Count = n(),
    Mean_Math = mean(math12),
    Sd_Math = sd(math12),
    Mean_Read = mean(read12),
    Sd_Read = sd(read12)
  ) |>
  round(2)
```

**Interpretation:**

There are 7,430 students in the sample. The math score averaged 52.13 with a standard deviation of 9.46. The reading score averaged 51.77 with a standard deviation of 9.41.

> (ii) Run the simple regression of `math12` on `read12` to obtain the OLS intercept and slope estimates. Report the results.

```{r}
#| label: c10_ii

c10m <- lm(math12 ~ read12, data = c10)

modelsummary(
  c10m,
  fmt = 2,
  statistic = NULL,
  gof_map = c("nobs", "r.squared"),
  coef_rename = c(
    "(Intercept)" = "Constant",
    "read12" = "Reading Score"
  )
)
```

> (iii) Does the intercept reported in part (ii) have a meaningful interpretation? Explain.

```{r}
#| label: c10_iii

ggplot(c10, aes(x = read12)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  theme_minimal()
```

**Interpretation:**

The intercept of 15.15 represents the predicted math score for a student with a reading score of exactly zero. This intercept does not have a meaningful real-world interpretation because a reading score of zero is far outside the range of our sample data (as shown in the histogram, the lowest scores are around 20). It is merely a mechanical extrapolation necessary to satisfy the OLS first-order conditions.

> (iv) Are you surprised by the $\hat \beta_1$ that you found? What about $R^2$?

**Interpretation:**

A high $\hat{\beta}_1$ is not surprising. Standardized tests for math and reading both heavily measure general cognitive ability, intelligence, and test-taking skills. Students who are academically gifted tend to perform well on all sections of a standardized test. Furthermore, an $R^2$ of around 0.50 (or 50%) is quite high for cross-sectional observational data on human behavior, confirming a strong linear relationship.

> (v) Suppose that you present your findings to a superintendent of a school district, and the superintendent says, “Your findings show that to improve math scores we just need to improve reading scores, so we should hire more reading tutors.” How would you respond to this comment? (Hint: If you instead run the regression of read12 on math12, what would you expect to find?)

```{r}
#| label: c10_v

c10m2 <- lm(read12 ~ math12, data = c10)

modelsummary(
  c10m2,
  fmt = 2,
  statistic = NULL,
  gof_map = c("nobs", "r.squared"),
  coef_rename = c(
    "(Intercept)" = "Constant",
    "math12" = "Math Score"
  )
)
```

**Interpretation:**

The superintendent is falsely equating correlation with causality. Running the regression in reverse yields the exact same $R^2$ because $R^2$ in a simple regression is just the squared correlation coefficient, which is symmetric. The relationship is driven by unobserved confounding variables (omitted variable bias)—such as a student's innate ability, family income, or the quality of the school itself—which simultaneously cause high reading scores and high math scores. Hiring reading tutors will likely only improve reading skills, not the underlying omitted variables driving the math scores.

::: callout-note
### Correlation VS Causaliy.

Remember that OLS slope can be written as

$$
\hat \beta_1 = \rho_{xy}\cdot \dfrac{\sigma_y}{\sigma_x}
$$

So it is measuring correlation at the end of the day, except if zero conditional mean assumption is really met.
:::

## C11 \[GPA1\]

>  Use the data in GPA1 to answer these questions. It is a sample of Michigan State University undergraduates from the mid-1990s, and includes current college GPA, `colGPA`, and a binary variable indicating whether the student owned a personal computer `PC`.

```{r}
#| label: c11_data

c11 <- gpa1
kable(head(c11))
```

> (i) How many students are in the sample? Find the average and highest college GPAs.

```{r}
#| label: c11_i

c11 |>
  summarise(
    Count = n(),
    Avg_GPA = mean(colGPA),
    Max_GPA = max(colGPA)
  ) |>
  round(2)
```

**Interpretation:**

There are 141 students in the sample. The average college GPA is 3.06, and the maximum GPA in the sample is 4.0.

> (ii) How many students owned their own PC?

```{r}
#| label: c11_ii

c11 |>
  summarise(
    Has_PC = sum(PC)
  )
```

**Interpretation:**

In this sample, 56 students own a personal computer.

> (iii) Estimate the simple regression equation $$colGPA = \beta_0 + PC + u$$ Interpret these estimates, including a discussion of the magnitudes.

```{r}
#| label: c11_iii

c11m <- lm(colGPA ~ PC, data = c11)

modelsummary(
  c11m,
  fmt = 2,
  statistic = NULL,
  gof_map = c("nobs", "r.squared"),
  coef_rename = c(
    "(Intercept)" = "Constant",
    "PC" = "Own PC"
  )
)
```

**Interpretation:**

The estimated equation shows that the predicted GPA for a student who does not own a PC is 2.99. Owning a PC is associated with a predicted GPA increase of 0.17 points, resulting in a total predicted GPA of 3.16. In terms of magnitude, a 0.17 point increase is practically significant, as it represents a meaningful jump in a student's letter grade (e.g., roughly half the distance between a B and a B+).

> (iv)  What is the R-squared from the regression? What do you make of its magnitude?

The R-squared is 0.05 (or 5%). This means that owning a PC explains only 5% of the variation in college GPAs. The magnitude is quite low, which makes logical sense; the vast majority (95%) of the variation in GPA is driven by other unobserved factors such as study habits, natural aptitude, major difficulty, and high school preparation. 

> (v) Does your finding in part (iii) imply that owning a PC has a causal effect on colGPA? Explain.

**Interpretation:**

No, this finding does not imply a causal effect. There is a severe risk of omitted variable bias. The binary variable for owning a PC is likely positively correlated with unobserved factors in the error term, such as family wealth, parental education, or the student's inherent motivation. For example, wealthier parents might be more likely to buy their child a PC, and wealth might also provide other educational advantages (like tutoring or not having to work a part-time job), which actually drive the higher GPA.

## C12 \[LOANAPP\]

> Use the data in LOANAPP to answer these questions. See also Computer Exercise C9 in Chapter 1. The variable `approve` is a binary variable set to one if an applicant’s mortgage application was approved and zero if not. The variable `white` is equal to one if the applicant is classified as being White and zero if not.

```{r}
#| label: c12_data

c12 <- loanapp
kable(head(c12))
```

> (i) How many mortgage applications are represented in the sample? What percentage of the applicants are classified as being White? Find the approval rates separately for `white = 0` and `white = 1`.

```{r}
#| label: c12_i

c12 |>
  summarise(
    Total_Applications = n(),
    Percent_White = mean(white) * 100
  ) |>
  round(2)
```

```{r}
#| label: c12_i_2

c12 |>
  summarise(
    Approval_Rate = round(mean(approve), 2),
    .by = white
  ) |>
  mutate(Is_White = ifelse(white == 1, "White", "Not White")) |>
  select(Is_White, Approval_Rate)
```

**Interpretation:**

The sample contains 1989 mortgage applications. Approximately 84.5% of the applicants in this sample are classified as White. The approval rate for White applicants is 91% (0.91), while the approval rate for non-White applicants is 71% (0.71).

> (ii) Run a simple regression of approve on white and report the results. Note that the OLS method allows `approve` and `white` to both be binary variables.

```{r}
#| label: c12_ii

c12m <- lm(approve ~ white, data = c12)

modelsummary(
  c12m,
  fmt = 2,
  statistic = NULL,
  gof_map = c("nobs", "r.squared"),
  coef_rename = c(
    "(Intercept)" = "Constant",
    "white" = "Is_White"
  )
)
```

> (iii) Interpret your estimate $\hat \beta_0$ in part(ii). How do you interpret $\hat \beta_1$? [Hint: See part(i).]

**Interpretation:**

The intercept ($\hat{\beta}_0$) is 0.71, which is the predicted probability of mortgage approval for a non-White applicant (white = 0). The slope ($\hat{\beta}_1$) is 0.20, meaning that being White is associated with a predicted 0.20 increase in the probability of approval, leading to a total predicted approval rate of 0.91 (or 91%) for White applicants.

> (iv)  Describe the variable `loanprc`. What are its average, standard deviation, minimum, and maximum values in the sample?

```{r}
#| label: c12_c

c12 |>
  summarise(
    LTP_Mean = mean(loanprc),
    LTP_Sd = sd(loanprc),
    LTP_Max = max(loanprc),
    LTP_Min = min(loanprc)
  ) |>
  round(2)
```

**Interpretation:**

The variable loanprc (the loan-to-price ratio) has a sample mean of 0.77 and a standard deviation of 0.19. The minimum loan-to-price ratio in the sample is 0.02 and it reaches a maximum of 2.57.

> (v)  Estimate a simple regression of `loanprc` on `white` and interpret the results.

```{r}
#| label: c12_v

c12m2 <- lm(loanprc ~ white, data = c12)

modelsummary(
  c12m2,
  fmt = 2,
  statistic = NULL,
  gof_map = c("nobs", "r.squared"),
  coef_rename = c(
    "(Intercept)" = "Constant",
    "white" = "Is_White"
  )
)
```

**Interpretation:**

The intercept indicates that the predicted loan-to-price ratio for a non-White applicant is 0.84. The slope indicates that White applicants have a predicted loan-to-price ratio that is 0.09 lower, resulting in a predicted ratio of 0.75.

> (vi) Does it make sense to think of `loanprc` as a variable that can affect the chance that a loan is approved? Explain.

**Interpretation:**

Yes, it makes perfect economic sense that loanprc affects approval chances. The loan-to-price ratio measures how much of the home's value is being financed versus how much is paid upfront as a down payment. A higher ratio means a smaller down payment, which translates to a higher risk of default for the bank. Therefore, a higher loanprc should naturally lower the probability of a loan being approved. Because loanprc fundamentally affects approval chances and is also correlated with race (as proven in part v), failing to include it in the regression causes severe omitted variable bias when attempting to measure racial discrimination in lending.
