---
title: "The Nature of Econometrics and Economic Data"
format: html
---

```{r}
#| label: 'setup'
#| message: false
#| warning: false

library(tidyverse)
library(wooldridge)
library(knitr)
library(janitor)
```

## C1 \[Wage 1\]

> Use the data in WAGE1 for this exercise.

```{r}
#| label: c1_data

c1 <- wage1
kable(head(c1))
```

> (i) Find the average education level in the sample. What are the lowest and highest years of education?

```{r}
#| label: c1_i

c1 |>
  summarise(
    Avg_Educ = mean(educ),
    Min_Educ = min(educ),
    Max_Educ = max(educ)
  ) |>
  round(2)
```

**Interpretation:**

The average education in this sample is approximately 12.5 years. The range is broad ranging from 0 years (no formal education) to 18 years of education (graduate level).

> (ii) Find the average hourly wage in the sample. Does it seem high or low?

```{r}
#| label: c1_ii

c1 |>
  summarise(
    Avg_Wage = mean(wage, na.rm = TRUE)
  ) |>
  round(2)
```

**Interpretation:**

The average nominal wage is \$5.90. On its own, in 1976 dollars, it might seem low compared to modern standards, but we need to account for inflation to truly judge its purchasing power.

> (iii) The wage data are reported in 1976 dollars. Using the Internet or a printed source, find the Consumer Price Index (CPI) for the years 1976 and 2022.

**Interpretation:**

CPI for 1976 is **\$56.9** and became **\$292.7** in 2022.

> (iv) Use the CPI values from part (iii) to find the average hourly wage in 2022 dollars. Now does the average hourly wage seem reasonable?

$$
\begin{align*}
{Wage}_{2022} &= {Wage}_{1976} \times \dfrac{CPI_{2022}}{CPI_{1976}}\\
&= 5.9 \times \dfrac{292.7}{56.9}\\
&= 30.35
\end{align*}
$$

**Interpretation:**

The average wage in this sample is equivalent to \$30.35 in 2022 dollars.The nominal 1976 average wage of \$5.90 is more than 2.5 times the 1976 minimum wage of \$2.30. Because the sample's average earnings vastly exceed the baseline wage of its own era, it is highly probable that this sample includes a significant proportion of skilled or experienced labor.

> (v) How many women are in the sample? How many men?

```{r}
#| label: c1_v

c1 |>
  summarise(
    Count = n(),
    Avg_Wage = mean(wage) |> round(2),
    .by = female
  ) |>
  mutate(Gender = ifelse(female == 1, "Female", "Male")) |>
  select(Gender, Count, Avg_Wage)
```

**Interpretation:**

Although sample is well balanced with 252 females and 274 males, there is a notable wage gap between the genders.

## C2 \[BWGHT\]

> Use the data in BWGHT to answer this question.

```{r}
#| label: c2_data

c2 <- bwght
kable(head(c2))
```

> (i) How many women are in the sample, and how many report smoking during pregnancy?

```{r}
#| label: c2_i

c2 |>
  summarise(
    Count = n(),
    Smokers = sum(cigs > 0),
    Non_Smokers = sum(cigs == 0)
  )
```

**Interpretation:**

The sample contains **1,388 women**. Out of these, **212 women** (approximately 15%) reported smoking at least one cigarette per day during pregnancy, while **1,176** reported not smoking at all.

> (ii) What is the average number of cigarettes smoked per day? Is the average a good measure of the “typical” woman in this case? Explain.

```{r}
#| label: c2_ii

c2 |>
  summarise(
    Avg_Cigs_All = mean(cigs)
  ) |>
  round(2)
```

```{r}
#| label: c2_ii_2

ggplot(c2, aes(x = cigs)) +
  geom_bar()
```

**Interpretation:**

The sample average is **2.09** cigarettes per day. However, this is not a good measure of a "typical" woman because the vast majority of pregnant females in the sample (about 85%) are non-smokers. The median is 0.

> (iii) Among women who smoked during pregnancy, what is the average number of cigarettes smoked per day? How does this compare with your answer from part (ii), and why?

```{r}
#| label: c2_iii

c2 |>
  summarise(
    Avg_Cigs_Smokers = mean(cigs[cigs > 0])
  ) |>
  round(2)
```

**Interpretation:**

A pregnant female who smokes consumes an average of **13.67** cigarettes per day. This is significantly higher than the sample average of 2.09 because this calculation excludes the overwhelming majority of women who bring the overall average down to near zero.

> (iv) Find the average of `fatheduc` in the sample. Why are only 1,192 observations used to compute this average?

```{r}
#| label: c2_iv

c2 |>
  summarise(
    Avg_Fath_Educ = mean(fatheduc, na.rm = TRUE),
    Missing_Obs = sum(is.na(fatheduc))
  ) |>
  round(2)
```

**Interpretation:**

An average father has **13.19** years of education in the sample, with *196* missing data.

> (v) Report the average family income and its standard deviation in dollars.

```{r}
#| label: c2_v

c2 |>
  summarise(
    Avg_Fam_Inc = mean(faminc, na.rm = TRUE),
    SD_Fam_Inc = sd(faminc, na.rm = TRUE)
  ) |>
  round(2)
```

**Interpretation:**

Families included in the sample have an average income of **\$29,000** with a standard deviation of **\$18,700**

## C3 \[MEAP01\]

> The data in MEAP01 are for the state of Michigan in the year 2001. Use these data to answer the following questions.

```{r}
#| label: c3_data

c3 <- meap01
kable(head(c3))
```

> (i) Find the largest and smallest values of `math4`. Does the range make sense? Explain.

```{r}
#| label: c3_i

c3 |>
  summarise(
    Max_math4 = max(math4),
    Min_math4 = min(math4)
  ) |>
  round(2)
```

**Interpretation:**

The sample includes math scores ranging from **0** to **100** indicating wide variety.

> (ii) How many schools have a perfect pass rate on the math test? What percentage is this of the total sample?

```{r}
#| label: c3_ii

c3 |>
  summarise(
    Total_Schools = n(),
    Perfect_Math_Count = sum(math4 == 100, na.rm = TRUE),
    Percentage = (Perfect_Math_Count / Total_Schools) * 100
  ) |>
  round(2)
```

**Interpretation:**

Out of the 1,823 schools in the sample, 38 schools (approximately 2.08%) achieved a perfect 100% pass rate on the fourth-grade math test.

> (iii) How many schools have math pass rates of exactly 50%?

```{r}
#| label: c3_iii

c3 |>
  summarise(
    Schools_at_50 = sum(math4 == 50, na.rm = TRUE)
  )
```

**Interpretation:**

There are exactly 17 schools with a pass rate of exactly 50%

> (iv) Compare the average pass rates for the math and reading scores. Which test is harder to pass?

```{r}
#| label: c3_iv

c3 |>
  drop_na(math4, read4) |>
  summarise(
    Avg_Math = mean(math4),
    Avg_Read = mean(read4)
  ) |>
  round(2)
```

**Interpretation:**

The average pass rate for math **71.90%** is significantly higher than for reading **60.06%**. This shows that the reading exam was harder in the sample.

> (v) Find the correlation between `math4` and `read4`. What do you conclude?

```{r}
#| label: c3_v

c3 |>
  drop_na(math4, read4) |>
  summarise(
    Correlation = cor(math4, read4)
  ) |>
  round(2)
```

**Interpretation:**

The correlation between the two exams is high **0.84** showing that performance in the two exams is similar.

> (vi) The variable `exppp` is expenditure per pupil. Find the average of `exppp` along with its standard deviation. Would you say there is wide variation in per pupil spending?

```{r}
#| label: c3_vi

c3 |>
  drop_na(math4, read4) |>
  summarise(
    Mean_Expnd = mean(exppp),
    Std_Expend = sd(exppp)
  ) |>
  round(2)
```

**Interpretation:**

The average expenditure per pupil is **\$5,194.87**, with a standard deviation of **\$1,091.89** representing a significant spread (**21%** of the mean).

::: callout-note
### Coefficient of Variation (CV)

To know if the spread is significant or not, calculate coefficient of variation (cv). $$
CV = \dfrac \sigma \mu$$

Small (cv \< 10%)

Medium (cv 10% - 30%)

Large(cv \> 30%)
:::

> (vii) Suppose School A spends \$6,000 per student and School B spends \$5,500 per student. By what percentage does School A’s spending exceed School B’s? Compare this to $100 \times [\log(6,000) – \log(5,500)]$, which is the approximation percentage difference based on the difference in the natural logs?

```{r}
#| label: c3_vii

school_a <- 6000
school_b <- 5500


pct_diff <- ((school_a - school_b) / school_b) * 100
log_diff <- 100 * (log(school_a) - log(school_b))


results <- c(Exact_Pct = pct_diff, Log_Approx = log_diff)
results |> round(2)
```

**Interpretation:**

School A spends **9.09%** more per student than School B. The logarithmic approximation ($100 \times \Delta \log$) yields **8.70**, which is a reasonably close estimate.

::: callout-important
### Logarithmic Approximation for Percentage Differences

When comparing two cross-sectional observations (such as expenditures between two different schools), we calculate a **percentage difference**. The exact percentage difference of $x_1$ relative to $x_0$ is:

$$\% \text{Difference} = \frac{x_1 - x_0}{x_0} \times 100$$

We can approximate this difference using natural logarithms:

$$\% \text{Difference} \approx 100 \times [\log(x_1) - \log(x_0)]$$

*Note:* This approximation is only mathematically reliable for small differences (typically less than 10%). For larger gaps, the exact formula must be used.
:::

## C4 \[JTRAIN2\]

> The data in JTRAIN2 come from a job training experiment conducted for low-income men during 1976–1977; see Lalonde (1986).

```{r}
#| label: c4_data

c4 <- jtrain2
kable(head(c4))
```

> (i) Use the indicator variable `train` to determine the fraction of men receiving job training.

```{r}
#| label: c4_i

c4 |>
  summarise(
    Trained = mean(train)
  ) |>
  round(2)
```

**Interpretation:**

**42%** of men in the sample received training.

> (ii) The variable `re78` is earnings from 1978, measured in thousands of 1982 dollars. Find the averages of `re78` for the sample of men receiving job training and the sample not receiving job training. Is the difference economically large?

```{r}
#| label: c4_ii

c4 |>
  summarise(
    Count = n(),
    Earnings = mean(re78) |> round(2),
    .by = train
  ) |>
  mutate(Train = ifelse(train == 1, "Trained", "Untrained")) |>
  select(Train, Count, Earnings)
```

**Interpretation:**

Trained men received earnings of **6,350** while untrained received **4,550** only. The percentage difference is $\%\Delta x =\dfrac{6.35 - 4.55}{4.55} =  39.56\%$ which is practically significant. A difference of 1800 is economically significant, especially for low-income men in 1978.

::: callout-caution
### Terminology: Change vs. Difference

**Change = Time (Longitudinal)** Tracking the *same* subject across different time periods. \* *Math:* (New - Old) / Old

**Difference = Distinct Groups (Cross-Sectional)** Comparing *separate* subjects at the *same* point in time. \* *Math:* (Group A - Group B) / Group B
:::

> (iii) The variable `unem78` is an indicator of whether a man was unemployed or not in 1978. What fraction of the men who received job training were unemployed? What about for men who did not receive job training? Comment on the difference.

```{r}
#| label: c4_iii

c4 |>
  summarise(
    Unemployed = mean(unem78) |> round(2),
    .by = train
  ) |>
  mutate(Train = ifelse(train == 1, "Trained", "Untrained")) |>
  select(Train, Unemployed)
```

**Interpretation:**

**24%** of the trained men were unemployed while **35%** of the untrained men were unemployed. There is a **11 percentage point difference** in unemployment between the two groups.

> (iv) From parts (ii) and (iii), does it appear that the job training program was effective? What would make our conclusions more convincing?

**Interpretation:**

Training program appears to be effective to increase earnings from part (ii) and decrease unemployment from part (iii). Conclusions will be more convincing if the random sample was balanced, and we controlled for other factors such as earnings and employment before taking the program.

## C5 \[FERTIL2\]

> The data in FERTIL2 were collected on women living in the Republic of Botswana in 1988. The variable `children` refers to the number of living children. The variable `electric` is a binary indicator equal to one if the woman’s home has electricity, and zero if not.

```{r}
#| label: c5_data

c5 <- fertil2
kable(head(c5))
```

> (i) Find the smallest and largest values of `children` in the sample. What is the average of `children`?

```{r}
#| label: c5_i

c5 |>
  summarise(
    Min = min(children),
    Max = max(children),
    Average = mean(children)
  ) |>
  round(2)
```

```{r}
#| label: c5_i_2

ggplot(c5, aes(x = factor(children))) +
  geom_bar(fill = "#2c3e50") +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5, size = 3, color = "black") +
  theme_minimal() +
  labs(
    title = "Distribution of Number of Children",
    subtitle = "Based on the c5 dataset observations",
    x = "Number of Children",
    y = "Frequency (Count)",
    caption = "Source: Wooldridge Dataset"
  )
```

**Interpretation**:

Number of children is heavily right skewed. **1132** women had 0 children while only one woman had a maximum number of **13** children.

> (ii) What percentage of women have electricity in the home?

```{r}
#| label: c5_ii

c5 |>
  summarise(
    Has_Electricity = mean(electric, na.rm = TRUE)
  ) |>
  round(2)
```

**Interpretation**:

**14%** of the women in the sample had home electricity.

> (iii) Compute the average of `children` for those without electricity and do the same for those with electricity. Comment on what you find.

```{r}
#| label: c5_iii

c5 |>
  drop_na(electric) |>
  summarise(
    Avg_Children = mean(children) |> round(2),
    .by = electric
  ) |>
  mutate(Home_Electricity = ifelse(electric == 1, "Has Electricity", "No Electricity")) |>
  select(Home_Electricity, Avg_Children)
```

**Interpretation:**

Women who had access to electricity have lower average children **(1.9)** compared to women with no electricity **(2.33)**

> (iv) From part (iii), can you infer that having electricity “causes” women to have fewer children? Explain.

**Interpretation:**

No, we cannot conclude that having electricity causes women to have fewer children based solely on this data. While there is a clear negative correlation, this simple comparison in averages suffers heavily from omitted variable bias. Electricity in the home is likely a proxy variable for broader confounding factors—such as higher household income, urban residency, or better access to education and healthcare—which are the more probable underlying drivers of lower fertility rates.

## C6 \[COUNTYMURDERS\]

> Use the data in COUNTYMURDERS to answer this question. Use only the year 1996. The variable `murders` is the number of murders reported in the county. The variable `execs` is the number of executions that took place of people sentenced to death in the given county. Most states in the United States have the death penalty, but several do not.

```{r}
#|label: c6_data

c6 <- countymurders |>
  filter(year == 1996)

kable(head(c6))
```

> (i) How many counties are there in the data set? Of these, how many have zero murders? What percentage of counties have zero executions?

```{r}
#| label: c6_i

c6 |>
  summarise(
    Counties = n_distinct(countyid),
    Zero_Murders = sum(murders == 0),
    Pct_Zero_Executions = mean(execs == 0) * 100
  ) |>
  round(2)
```

**Interpretation:**

There are **2197** counties in the sample. Out of these, **1051** have zero murders, and **98.59%** of them have zero executions.

> (ii) What is the largest number of murders? What is the largest number of executions? Compute the average number of executions and explain why it is so small.

```{r}
#| label: c6_ii

c6 |>
  summarise(
    Max_Murders = max(murders),
    Max_Execs = max(execs),
    Avg_Execs = mean(execs)
  ) |>
  round(2)
```

```{r}
#| label: c6_ii_2

c6 |>
  tabyl(execs) |>
  adorn_pct_formatting()
```

**Interpretation:**

Largest number of murders is **1403** while largest number of executions is only **3**. Average execution is **0.02** because 98.6% of the sample have zero executions, 1.3% have 1 execution, leaving only 0.1% for 2 executions.

> (iii) Compute the correlation coefficient between `murders` and `execs` and describe what you find.

```{r}
#| label: c6_iii

c6 |>
  summarise(
    Correlation = cor(murders, execs)
  ) |>
  round(2)
```

**Interpretation:**

There is a weak correlation **(0.21)** between murders and executions.

> (iv) You should have computed a positive correlation in part (iii). Do you think that more executions `cause` more murders to occur? What might explain the positive correlation?

**Interpretation:**

No, more executions do not "cause" more murders. The positive, albeit weak, correlation (0.21) is highly misleading for several reasons:

-   **Omitted Variable Bias:** The raw counts ignore population size. Highly populated counties will naturally have higher absolute numbers of both murders and executions compared to rural counties.
-   **Reverse Causality:** The causal arrow likely points the other direction. Counties experiencing higher murder rates might face intense public and political pressure to utilize the death penalty more frequently in response to the crime wave.
-   **Justice System Lag & Attrition:** Executions in 1996 are punishments for crimes committed years or decades prior. Furthermore, many murders go unsolved, meaning the immediate murder count does not directly map to execution rates.
-   **Extreme Data Skew (Zero-Inflation):** Because 98.6% of the counties had exactly zero executions, this data is heavily right-skewed and zero-inflated. A standard linear correlation metric is mathematically inadequate for capturing the true dynamic here.

## C7 \[ALCOHOL\]

> The data set in ALCOHOL contains information on a sample of men in the United States. Two key variables are self-reported employment status and alcohol abuse (along with many other variables). The variables `employ` and `abuse` are both binary, or indicator, variables: they take on only the values zero and one.

```{r}
#| label: c7_data

c7 <- alcohol
kable(head(c7))
```

> (i) What percentage of the men in the sample report abusing alcohol? What is the employment rate?

```{r}
#| label: c7_i

c7 |>
  summarise(
    Pct_Alcohol_Abuse = mean(abuse) * 100,
    Employment_Rate = mean(employ)
  ) |>
  round(3)
```

**Interpretation:**

**9.917%** of men in the sample reported abusing alcohol. Self reported employment rate is **0.898** so 89.8% of the men are employed.

> (ii) Consider the group of men who abuse alcohol. What is the employment rate?

```{r}
#| label: c7_ii

c7 |>
  filter(abuse == 1) |>
  summarise(
    Count = n(),
    Employment_Rate = mean(employ)
  ) |>
  round(3)
```

**Interpretation:**

Employment rate for alcohol abusers is **0.873**.

> (iii) What is the employment rate for the group of men who do not abuse alcohol?

```{r}
#| label: c7_iii

c7 |>
  filter(abuse == 0) |>
  summarise(
    Count = n(),
    Employment_Rate = mean(employ)
  ) |>
  round(3)
```

**Interpretation:**

For non alcohol abusers, employment rate is 0.901.

> (iv) Discuss the difference in your answers to parts (ii) and (iii). Does this allow you to conclude that alcohol abuse causes unemployment?

**Interpretation:**

No, we cannot conclude that alcohol abuse causes unemployment. There is a 3 percentage point difference in the employment rates (87% for abusers versus 90% for non-abusers). However, this is merely a correlation. This simple comparison fails to account for omitted variable bias. Other unobserved factors may be education level, underlying mental or physical health issues.

::: callout-note
### Random Sampling vs. Over-Sampling

A true **random sample** naturally reflects the base rates of the actual population. If only 10% of the real world abuses alcohol, a perfectly random sample `must` show a roughly 9-to-1 imbalance.

**Over-sampling** happens when researchers intentionally skew the collection process to gather more data on a minority group (e.g., artificially forcing a 50/50 split).

Therefore, massively unequal group sizes in your data are a sign of healthy randomization.
:::

## C8 \[ECONMATH\]

> The data in ECONMATH were obtained on students from a large university course in introductory microeconomics. For this problem, we are interested in two variables: `score`, which is the final course score, and `econhs`, which is a binary variable indicating whether a student took an economics course in high school

```{r}
#| label: c8_data

c8 <- econmath
kable(head(c8))
```

> (i) How many students are in the sample? How many students report taking an economics course in high school?

```{r}
#| label: c8_i

c8 |>
  summarise(
    Count = n(),
    HS_Economics = sum(econhs)
  )
```

**Interpretation:** The sample has **856** students. Among those students, **317** took the economics class in high school.

> (ii) Find the average of score for those students who did take a high school economics class. How does it compare with the average of score for those who did not?

```{r}
#| label: c8_ii
c8 |>
  summarise(
    Avg_Score = round(mean(score), 2),
    .by = econhs
  ) |>
  mutate(HS_Economics = ifelse(econhs == 1, "HS Economics", "No HS Economics ")) |>
  select(HS_Economics, Avg_Score)
```

**Interpretation:** Average score for students with no high school economics is **72.91** while students with high school economics scored on average **72.08**

> (iii) Do the findings in part (ii) necessarily tell you anything about the causal effect of taking high school economics on college course performance? Explain.

No, students with high school economics chose to take the course in high school so we have selection bias. There is a small difference between the two groups that may be due to other factors or pure chance.

> (iv) If you want to obtain a good causal estimate of the effect of taking a high school economics course using the difference in averages, what experiment would you run?

**Interpretation:** I would conduct an **RCT** in which I randomly assign students to high school economics and other students stay as control. Then I will track both groups score in the university course.

## C9 \[LOANAPP\]

> The data set in LOANAPP contains information on a sample of mortgage applications in the Boston, Massachusetts area; the data were originally compiled by the Federal Reserve Bank of Boston. The race variable, denoted `white`, is recorded as binary, equaling one if the mortgage applicant is classified as White and zero otherwise. The variable `approve` is also binary, equaling one if the loan application was approved, and zero otherwise.

```{r}
#| label: c9_data

c9 <- loanapp
kable(head(c9))
```

> (i) What proportion of the applicants in the sample are recorded as being White?

```{r}
#| label: c9_i

c9 |>
  summarise(
    Pct_White = mean(white) * 100
  ) |>
  round(2)
```

**Interpretation:** **84.51%** of the applicants in the sample are white.

> (ii) For the group of non-White applicants, what is the proportion of loans approved?

```{r}
#| label: c9_ii

c9 |>
  filter(white == 0) |>
  summarise(
    Approved_Pct = mean(approve) * 100
  ) |>
  round(2)
```

**Interpretation:**

Among non-white applicants, **70.78%** have their application approved.

> (iii) What is the loan approval rate for White applicants?

```{r}
#| label: c9_iii

c9 |>
  filter(white == 1) |>
  summarise(
    Approved_Pct = mean(approve) * 100
  ) |>
  round(2)
```

**Interpretation:**

Among white applicants, **90.84%** have their application approved.

> (iv) Report the difference between (iii) and (ii) in percentage point terms. Does this analysis allow you to convincingly conclude that there is racial bias in mortgage application approvals? Explain.

**Interpretation:**

The percentage point difference is $90.84 - 70.78 = 20.06$. Although the gap between both groups is big, it can be due to other factors like employment status and crime record.

::: callout-warning
### Percentage Difference vs. Percentage Point Difference

**Percentage Point Difference:** The absolute mathematical difference between two rates.

$$\text{Difference in Points} = \text{Rate}_A - \text{Rate}_B$$

**Percentage Difference:** The relative size of the gap compared to a chosen base group.

$$\% \text{Difference} = \frac{\text{Rate}_A - \text{Rate}_B}{\text{Rate}_B} \times 100$$
:::
