---
title: "Multiple Regression Analysis: Further Issues"
format: html
---

## Computer Exercises

```{r}
#| label: 'setup'
#| include: false
library(tidyverse)
library(wooldridge)
library(modelsummary)
library(car)
library(broom)
library(ggcorrplot)
```

### C1 \[KIELMC\]

```{r}
df_1981 <- kielmc |>
  filter(year == 1981)
knitr::kable(head(df_1981))
```

> (i)

```{r}
#| label: c1_i
model_c1_i <- lm(lprice ~ ldist, data = df_1981)

modelsummary(
  model_c1_i,
  fmt = 2,
  shape = term ~ model + statistic,
  estimate = "{estimate}{stars}",
  statistic = c("std.error", "statistic", "p.value"),
  gof_map = c("nobs", "r.squared", "adj.r.squared"),
  coef_rename = c(
    "(Intercept)" = "Constant",
    "ldist" = "Distance from Incinerator"
  )
)
```

**Interpretation:**

Incinerator omits wastes that should reduce house prices. $\beta_1$ is expected to have a positive effect. $\beta_1$ = 0.36, for each **1%** increase in distance from incinerator, prices are expected to increase by **0.36%**. The result is statistically significant supporting the theory that proximity to the incinerator negatively affects property value. The estimate may be biased if its correlated with other omitted factors.

> (ii)

```{r}
#| label: c1_ii
model_c1_ii <- lm(
  lprice ~ ldist + lintst + larea + lland +
    rooms + baths + age,
  data = df_1981
)

modelsummary(
  model_c1_ii,
  fmt = 2,
  shape = term ~ model + statistic,
  estimate = "{estimate}{stars}",
  statistic = c("std.error", "statistic", "p.value"),
  gof_map = c("nobs", "r.squared", "adj.r.squared"),
  coef_rename = c(
    "(Intercept)" = "Constant",
    "ldist" = "Log Dist. to Incinerator",
    "lintst" = "Log Dist. to Interstate",
    "larea" = "Log Square Footage of House",
    "lland" = "Log Lot size",
    "rooms" = "Number of Rooms",
    "baths" = "Number of Bathrooms",
    "age" = "Age of House in Years"
  )
)
```

```{r}
#| label: c1_ii_2
#| warning: false
#| message: false
df_1981_sub <- df_1981 |>
  select(c("ldist", "lintst", "lprice", "age", "rooms", "larea", "lland", "baths"))

corr_matrix <- cor(df_1981_sub, use = "complete.obs")

ggcorrplot(corr_matrix,
  hc.order = TRUE,
  type = "lower",
  lab = TRUE,
  title = "Correlation Matrix (1981 Housing Data)"
)
```

**Interpretation:**

In the multiple regression model, the coefficient for distance to the incinerator ($\beta_1$) shrinks considerably and becomes statistically insignificant. The large, significant estimate observed in the simple model was heavily driven by omitted variable bias; distance from the incinerator is positively correlated with other desirable, price-increasing characteristics, such as larger lot sizes and closer proximity to the interstate. By including these controls, the isolated effect of the incinerator is revealed to be much smaller. Furthermore, the fact that these newly added spatial variables also lack individual statistical significance strongly suggests multicollinearity. The model struggles to disentangle the individual marginal effects of these highly correlated geographic features, resulting in inflated standard errors.

> (iii)

```{r}
#| label: c1_iii
model_c1_iii <- lm(
  lprice ~ ldist + lintst + I(lintst^2) + larea + lland +
    rooms + baths + age,
  data = df_1981
)

modelsummary(
  model_c1_iii,
  fmt = 2,
  shape = term ~ model + statistic,
  estimate = "{estimate}{stars}",
  statistic = c("std.error", "statistic", "p.value"),
  gof_map = c("nobs", "r.squared", "adj.r.squared"),
  coef_rename = c(
    "(Intercept)" = "Constant",
    "ldist"       = "Log Dist. to Incinerator",
    "lintst"      = "Log Dist. to Interstate",
    "I(lintst^2)" = "Log Dist. to Interstate Squared",
    "larea"       = "Log Square Footage of House",
    "lland"       = "Log Lot size",
    "rooms"       = "Number of Rooms",
    "baths"       = "Number of Bathrooms",
    "age"         = "Age of House in Years"
  )
)
```

```{r}
#| label: c1_iii_2
df_1981 |>
  summarize(
    Correlation = cor(ldist, lintstsq)
  )
```

**Interpretation:**

By adding the squared term for the distance to the interstate, we corrected a functional form misspecification present in the previous model. Both the linear term (2.07) and the squared term (-0.12) are statistically significant at the 1% level. The negative coefficient on the squared term indicates an inverted U-shaped relationship: moving away from the interstate initially increases house prices (escaping the noise penalty), but eventually, the marginal effect becomes negative as the house becomes too isolated (the commute penalty).With the non-linear spatial effect of the interstate properly modeled, the isolated effect of the incinerator is clarified. The coefficient $\beta_1 = 0.19$ is statistically significant ($p < 0.01$). Holding all other controlled factors constant, a 1% increase in distance from the incinerator is associated with a 0.19% increase in the housing price.

::: callout-note
Optimal distance from interstate is obtained by solving

$$
\frac{\partial \log(price)}{\partial \log(intst)} = 2.07 - 2(0.12) \log(intst) = 0
$$

Which can be solved easily using the formula

$$
x^* = \left| \dfrac{\hat \beta_1}{2 \hat \beta_2}\right| = \dfrac{2.07}{0.24} = 8.625
$$

Where $x^*$ is in log terms, so $\exp{x^*} =  \exp(8.625) = 5569$
:::

> (iv)

```{r}
#| label: c1_iv
model_c1_iv <- lm(
  lprice ~ ldist + I(ldist^2) + lintst + I(lintst^2)
    + larea + lland + rooms + baths + age,
  data = df_1981
)

modelsummary(
  model_c1_iv,
  fmt = 2,
  shape = term ~ model + statistic,
  estimate = "{estimate}{stars}",
  statistic = c("std.error", "statistic", "p.value"),
  gof_map = c("nobs", "r.squared", "adj.r.squared"),
  coef_rename = c(
    "(Intercept)" = "Constant",
    "ldist"       = "Log Dist. to Incinerator",
    "I(ldist^2)"  = "Log Dist. to Incinerator squared",
    "lintst"      = "Log Dist. to Interstate",
    "I(lintst^2)" = "Log Dist. to Interstate Squared",
    "larea"       = "Log Square Footage of House",
    "lland"       = "Log Lot size",
    "rooms"       = "Number of Rooms",
    "baths"       = "Number of Bathrooms",
    "age"         = "Age of House in Years"
  )
)
```

**Interpretation:**

Adding the squared term for the distance to the incinerator causes both the linear ($\beta_1 = 0.87$, $p = 0.67$) and quadratic ($-0.04$, $p = 0.74$) terms to become statistically insignificant. This insignificance is driven by severe variance inflation due to structural multicollinearity between $\log(dist)$ and $[\log(dist)]^2$. We conclude that there is no evidence of a non-linear relationship between house prices and proximity to the incinerator. The strictly linear functional form for $\log(dist)$ established in part (iii) is the correct specification.

### C2 \[WAGE1\]

```{r}
#| label: c2_data
df_wage1 <- wooldridge::wage1
knitr::kable(head(df_wage1))
```

> (i)

```{r}
#| label: c2_i
model_c2 <- lm(
  lwage ~ educ + exper + I(exper^2),
  data = df_wage1
)

modelsummary(
  model_c2,
  fmt = 6,
  shape = term ~ model + statistic,
  estimate = "{estimate}{stars}",
  statistic = c("std.error", "statistic", "p.value"),
  gof_map = c("nobs", "r.squared", "adj.r.squared"),
  coef_rename = c(
    "(Intercept)" = "Constant",
    "educ"        = "Education",
    "exper"       = "Experience",
    "I(exper^2)"  = "Experience Squared"
  )
)
```

> (ii)

**Interpretation:** Yes, although the estimated effect is $-0.001$ which is almost zero, the standard error is microscopic.

::: callout-important
Think of what experience squared represents, a 30 years of experience will be 900 years.
:::

> (iii)

**The Math Way:** 
$$\begin{align}
100(0.041 + 2(-0.001)\cdot 4) &= 3.53\%\\
100(0.041 + 2(-0.001) \cdot 19 &= 1.3\%)
\end{align}$$


**The Code Way:**

```{r}
#| label: c2_iii
compare_df <- data.frame(
  exper = c(4, 5),
  educ = 12
)

log_wages <- predict(model_c2, newdata = compare_df)
log_diff <- log_wages[2] - log_wages[1]

pct_change_exact <- 100 * (exp(log_diff) - 1)
pct_change_approx <- 100 * log_diff

paste(pct_change_approx, pct_change_exact)
```

::: callout-note
Why do my returns differ (3.53% vs. 3.45% vs. 3.52%)?

The discrepancy comes from how we handle the "curvature" of the Experience Squared term:

1.  **The Math Way (3.53%):** Uses the **derivative** ($100 \cdot [\beta_2 + 2\beta_3 \cdot 4]$). This is a linear approximation. It assumes the slope is a flat line at exactly Year 4. It ignores the "bend" that happens as you move toward Year 5.
2.  **Approx Predict (3.46%):** This is the **Log Difference** ($\hat{y}_5 - \hat{y}_4$). It accounts for the curve between the two years but stays in "log points."
3.  **Exact Predict (3.52%):** This is the **Exponential Change** ($100 \cdot [e^{(\hat{y}_5 - \hat{y}_4)} - 1]$). This is the only mathematically "perfect" representation of the percentage increase in actual dollars.
:::

**Interpretation:**

The estimated return to experience is positive but diminishing. Obtaining a fifth year of experience (moving from 4 to 5 years) is associated with an expected wage increase of approximately 3.5%, holding other factors constant. However, by the twentieth year of experience (moving from 19 to 20 years), the marginal return drops to approximately 1.3%

> (iv)

$$
\begin{align}
\% \Delta \widehat{wage} &\approx 100(\hat \beta_2 + 2 \hat \beta_3 exper) \Delta exper \\
0 &= 100(\hat \beta_2 + 2 \hat \beta_3 exper) \Delta exper\\
0 &= \hat \beta_2 + 2 \hat \beta_3 exper\\
2 \hat \beta_3 exper &= - \hat \beta_2\\
exper &= \dfrac{- \hat \beta_2}{2 \hat \beta_3}\\
exper &= \dfrac{-0.04101}{2(-0.00071)} \\
&\approx 28.88 \text{ years}
\end{align}
$$

```{r}
#| label: c2_iv
df_wage1 |>
  summarise(
    n_declining = sum(exper >= 29),
    pct_declining = mean(exper >= 29) * 100
  ) |>
  round(2)
```

**Interpretation:**

The optimal amount of experience for maximizing wages is approximately 29 years. Then the return becomes negative. The sample includes 121 workers with declining return of experience. These workers account for 23% of the workers in the sample.

### C3 [WAGE2]
