---
title: "The Nature of Econometrics and Economic Data"
format: html
---

```{r}
#| label: 'setup'
library(tidyverse)
library(wooldridge)
```

## C1 \[Wage 1\]

> Use the data in WAGE1 for this exercise.

```{r}
#| label: c1_data
c1 <- wooldridge::wage1
knitr::kable(head(c1))

```

> (i) Find the average education level in the sample. What are the lowest and highest years of education?

```{r}
#| label: c1-i
c1 |>
  summarise(
    Avg_Educ = mean(educ),
    Min_Educ = min(educ),
    Max_Educ = max(educ)
  ) |>
  round(2)
```

**Interpretation:**\
The average education in this sample is approximately 12.5 years. The range is broad ranging from 0 years (no formal education) to 18 years of education (graduate level)

> (ii) Find the average hourly wage in the sample. Does it seem high or low?

```{r}
#| label: c1-ii
c1 |>
  summarise(
    Avg_Wage = mean(wage, na.rm = TRUE)
  ) |>
  round(2)

```

$$
\begin{align*}
{Wage}_{2026} &= {Wage}_{1976} \times \dfrac{CPI_{2026}}{CPI_{1976}}\\
&= 5.9 \times \dfrac{325.3}{56.9}\\
&= 33.7
\end{align*}
$$

**Interpretation:**\
The average nominal wage of **\$5.90** in 1976 translates to approximately **\$33.73** in 2026 dollars. This adjustment reveals that the workers in the `wage1` dataset were significantly better off than the nominal figure suggests, earning a real wage that is competitive even by modern standards.

> (iii) The wage data are reported in 1976 dollars. Using the Internet or a printed source, find the Consumer Price Index (CPI) for the years 1976 and 2013.

**Interpretation:**\
CPI for 1976 is **\$56.9** and became **\$233** in 2013.

> (iv) Use the CPI values from part (iii) to find the average hourly wage in 2013 dollars. Now does the average hourly wage seem reasonable?

$$
\begin{align*}
{Wage}_{2013} &= {Wage}_{1976} \times \dfrac{CPI_{2013}}{CPI_{1976}}\\
&= 5.9 \times \dfrac{233}{56.9}\\
&= 24.2
\end{align*}
$$ **Interpretation:**\
The average wage is equivalent to **\$24.2** in 2013, which is three times the minimum wage of **\$7.25**. Minimum wage was **\$2.3** in 1976 which indicates sample includes skilled labor.

> (v) How many women are in the sample? How many men?

```{r}
#| label: c1-iv
c1 |>
  summarise(
    Count = n(),
    Avg_Wage = mean(wage) |> round(2),
    .by = female
  ) |>
  mutate(Gender = ifelse(female == 1, "Female", "Male")) |>
  select(Gender, Count, Avg_Wage)
```

**Interpretation:**\
Although sample is well balanced with 252 females and 274 males, there is a notable wage gap between the genders.

## C2 \[BWGHT\]

> Use the data in BWGHT to answer this question.

```{r}
#| label: c2_data
c2 <- wooldridge::bwght
knitr::kable(head(c2))

```

> (i) How many women are in the sample, and how many report smoking during pregnancy?

```{r}
#| label: c2_i
c2 |>
  summarise(
    Count = n(),
    Smokers = sum(cigs > 0),
    Non_Smokers = sum(cigs == 0)
  )
```

**Interpretation:**\
The sample contains **1,388 women**. Out of these, **212 women** (approximately 15%) reported smoking at least one cigarette per day during pregnancy, while **1,176** reported not smoking at all.

> (ii) What is the average number of cigarettes smoked per day? Is the average a good measure of the “typical” woman in this case? Explain.

```{r}
#| label: c2_ii
c2 |>
  summarise(
    Avg_Cigs_All = mean(cigs),
    Avg_Cigs_Smokers = mean(cigs[cigs > 0])
  ) |>
  round(2)
```

**Interpretation:**\
The sample average is **2.09** because most of the pregnant females are non smokers. A smoker pregnant female smokes on average **13.67** cigarettes per day.

> (iii) Among women who smoked during pregnancy, what is the average number of cigarettes smoked per day? How does this compare with your answer from part (ii), and why?

**Interpretation:**\
A smoker pregnant female smokes on average **13.67** cigarettes per day.

> (iv) Find the average of *fatheduc* in the sample. Why are only 1,192 observations used to compute this average?

```{r}
#| label: c2_iv
c2 |>
  summarise(
    Avg_Fath_Educ = mean(fatheduc, na.rm = TRUE),
    Missing_Obs = sum(is.na(fatheduc))
  ) |>
  round(2)
```

**Interpretation:**\
An average father has **13.19** years of education in the sample, with *196* missing data.

> (v) Report the average family income and its standard deviation in dollars.

```{r}
#| label: c2_v
c2 |>
  summarise(
    Avg_Fam_Inc = mean(faminc, na.rm = TRUE),
    SD_Fam_Inc = sd(faminc, na.rm = TRUE)
  ) |>
  round(2)
```

**Interpretation:**\
Families included in the sample have an average income of **\$29,000** with a standard deviation of **\$18,700**

## C3 \[MEAP01\]

> The data in MEAP01 are for the state of Michigan in the year 2001. Use these data to answer the following questions.

```{r}
#| label: c3_data
c3 <- wooldridge::meap01
knitr::kable(head(c3))

```

> (i) Find the largest and smallest values of *math4*. Does the range make sense? Explain.

```{r}
#| label: c3_i
c3 |>
  summarise(
    Max_math4 = max(math4),
    Min_math4 = min(math4)
  ) |>
  round(2)
```

**Interpretation:**\
The sample includes math scores ranging from **0** to **100** indication wide variety.

> (ii) How many schools have a perfect pass rate on the math test? What percentage is this of the total sample?

```{r}
#| label: c3_ii
c3 |>
  summarise(
    Total_Schools = n(),
    Perfect_Math_Count = sum(math4 == 100, na.rm = TRUE),
    Percentage = (Perfect_Math_Count / Total_Schools) * 100) |>
  round(2)

```

**Interpretation:**\
Out of the 1,823 schools in the sample, 38 schools (approximately 2.08%) achieved a perfect 100% pass rate on the fourth-grade math test.

> (iii) How many schools have math pass rates of exactly 50%?

```{r}
#| label: c3_iii
c3 |>
  summarise(
    Schools_at_50 = sum(math4 == 50, na.rm = TRUE)
)

```

**Interpretation:**\
There are exactly 17 schools with a pass rate of exactly 50%

> (iv) Compare the average pass rates for the math and reading scores. Which test is harder to pass?

```{r}
#| label: c3_iv
c3 |>
  drop_na(math4, read4) |> 
  summarise(
    Avg_Math = mean(math4),
    Avg_Read = mean(read4)
  ) |>
  round(2)
```

**Interpretation:**\
The average pass rate for math **71.90%** is significantly higher than for reading **60.06%**. This shows that the reading exam was harder in the sample.

> (v) Find the correlation between *math4* and *read4*. What do you conclude?

```{r}
#| label: c3_v
c3 |>
  drop_na(math4, read4) |> 
  summarise(
    Correlation = cor(math4, read4)
  ) |>
  round(2)

```

**Interpretation:**\
The correlation between the two exams is high **0.84** showing that performance in the two exams is similar.

> (vi) The variable *exppp* is expenditure per pupil. Find the average of *exppp* along with its standard deviation. Would you say there is wide variation in per pupil spending?

```{r}
#| label: c3_vi
c3 |>
  drop_na(math4, read4) |> 
  summarise(
    Mean_Expnd = mean(exppp),
    Std_Expend = sd(exppp)
  ) |>
  round(2)

```

**Interpretation:**\
The average expenditure per pupil is **\$5,194.87**, with a standard deviation of **\$1,091.89** representing a significant spread (**21%** of the mean).

::: callout-note
### Coefficient of Varitation (CV)

To know if the spread is significant or not, calculate coefficient of variation (cv). $$
CV = \dfrac \sigma \mu$$

Small (cv \< 10%)

Medium (cv 10% - 30%)

Large(cv \> 30%), mean is not reliable here.
:::

> (vii) Suppose School A spends \$6,000 per student and School B spends \$5,500 per student. By what percentage does School A’s spending exceed School B’s? Compare this to $100 \times [\log(6,000) – \log(5,500)]$, which is the approximation percentage difference based on the difference in the natural logs?

```{r}
#| label: c3-vii

school_a <- 6000
school_b <- 5500


pct_diff <- ((school_a - school_b) / school_b) * 100
log_diff <- 100 * (log(school_a) - log(school_b))


results <- c(Exact_Pct = pct_diff, Log_Approx = log_diff)
results |> round(2)
```

**Interpretation:**\
School A spends **9.09%** more per student than School B. The logarithmic approximation ($100 \times \Delta \log$) yields **8.70**, which is a reasonably close estimate.

::: callout-important
### Logarithmic Approximation

we can approximate **percentage change** $$\% \Delta x =\dfrac{x_1 - x_0}{x_0}$$ by $$100 \times [\log(x_1) - \log(x_0)]$$ Logarithmic approximation fails when percentage change is large (\> 10%)
:::

## C4 \[JTRAIN2\]

> The data in JTRAIN2 come from a job training experiment conducted for low-income men during 1976–1977; see Lalonde (1986).

```{r}
#| label: c4_data
c4 <- wooldridge::jtrain2
knitr::kable(head(c4))
```

> (i) Use the indicator variable *train* to determine the fraction of men receiving job training.

```{r}
#| label: c4_i
c4 |>
  summarise(
    Trained = mean(train)
  ) |>
  round(2)
```

**Interpretation:** **42%** of men in the sample received training.

> (ii) The variable *re78* is earnings from 1978, measured in thousands of 1982 dollars. Find the averages of *re78* for the sample of men receiving job training and the sample not receiving job training. Is the difference economically large?

```{r}
#| label: c4_ii
c4 |>
  summarise(
    Count = n(),
    Earnings = mean(re78) |> round(2),
    .by = train
  ) |>
  mutate(Train = ifelse(train == 1, "Trained", "Untrained")) |>
  select(Train, Count, Earnings)
```

**Interpretation:**
Trained men received earnings of **6.35** while untrained received **4.55* only. The percentage change is 
$\%\Delta x\dfrac{6.35 - 4.55}{4.55} =  39.56\%$ which is practically significant.

> (iii) The variable *unem78* is an indicator of whether a man was unemployed or not in 1978. What fraction of the men who received job training were unemployed? What about for men who did not receive job training? Comment on the difference.

```{r}
#| label: c4_iii
c4 |>
  summarise(
    Unemployed = mean(unem78) |> round(2),
    .by = train
  ) |>
  mutate(Train = ifelse(train == 1, "Trained", "Untrained")) |>
  select(Train, Unemployed)
```

**Interpretation:**
**24%** of the trained men where unemployed while **35%** of the untrained men where unemployed. There is a **11 percentage point change** in unemployment between the two groups.

> (iv) From parts (ii) and (iii), does it appear that the job training program was effective? What would make our conclusions more convincing?

**Interpretation:**
Training program appears to be effective to increase earnings from part (ii) and decrease unemployment from part (iii). Conclusions will be more convincing if the random sample was balanced, and we controlled for other factors such as earnings and employment before taking the program.