---
title: "The Simple Regression Model"
format: html
---

```{r}
#| label: 'setup'
#| message: false
#| warning: false

library(tidyverse)
library(wooldridge)
library(knitr)
library(modelsummary)
library(broom)
```

## C1 \[401K\]

```{r}
#| label: c1_data

c1 <- k401k
kable(head(c1))
```

> (i)

```{r}
#| label: c1_i

c1 |>
  summarise(
    Avg_Part_Rt = mean(prate),
    Avg_Match_Rt = mean(mrate)
  ) |>
  round(2)
```

**Interpretation:**

On average, **87.36%** of eligible employees participate in their company's 401(k) plan. Their company matches *\$0.73* for every dollar an employee contributes.

> (ii)

```{r}
#| label: c1_ii

c1m <- lm(prate ~ mrate, data = c1)

modelsummary(
  c1m,
  fmt = 2,
  statistic = NULL,
  gof_map = c("nobs", "r.squared"),
  coef_rename = c("mrate" = "Match Rate ($)", "(Intercept)" = "Constant")
)
```

**Interpretation:**

$$
\begin{align}
\beta_0 &= 83.03 & \beta_1 &= 5.86 \\
R^2 &= 0.075 & n &= 1534
\end{align}
$$

> (iii)

**Interpretation:**

If the match rate is 0, predicted participation rate is **83.08%**. For ever \$1 increase in match rate, participation is associated by an increase of **5.86 percentage points**.

> (iv)

```{r}
#| label: c1_iv

new_data <- data.frame(mrate = 3.5)

c1m |>
  predict(newdata = new_data) |>
  round(2)
```

::: callout-note
### Level-Level model prediction

To predict $\hat y$ value given a new value of x in a level-level model, substitute in this formula

$$\hat y = \hat \beta_0 + \hat \beta_1 x$$
:::

**Interpretation:**

Model predicts a participation rate of **103.59** which exceeds the maximum possible value of 100.

> (v)

**Interpretation:**

Match rate explains only about **7.5%** of the variation in participation rate, while the remaining 92.5% are due to other factors.

## C2 \[CEOSAL2\]

```{r}
#| label: c2_data

c2 <- ceosal2
kable(head(c2))
```

> (i)

```{r}
#| label: c2_i

c2 |>
  summarise(
    Avg_Salary = mean(salary),
    Avg_Tenure = mean(ceoten)
  ) |>
  round(2)
```

**Interpretation:**

Average salary is **\$865,860** with an average tenure of **7.95** years.

> (ii)

```{r}
#| label: c2_ii

c2 |>
  summarise(
    Fresh_CEO = sum(ceoten == 0),
    Veteran = max(ceoten)
  )
```

**Interpretation:**

There are **5** CEOs in their first year, and the longest serving CEO has been in charge for **37** years.

> (iii)

```{r}
#| label: c2_iii

c2m <- lm(lsalary ~ ceoten, data = c2)

modelsummary(
  c2m,
  fmt = 2,
  statistic = NULL,
  gof_map = c("nobs", "r.squared"),
  coef_rename = c("ceoten" = "Tenure", "(Intercept)" = "Constant")
)
```

**Interpretation:**

For every 1 additional year of tenure as CEO, salary is predicted to increase by **1%**. This shows that salary increases due to other factors not loyalty.

## C3 \[SLEEP75\]

```{r}
#| label: c3_data

c3 <- sleep75
kable(head(c3))
```

> (i)

```{r}
#| label: c3_i

c3m <- lm(sleep ~ totwrk, data = c3)

modelsummary(
  c3m,
  fmt = 2,
  statistic = NULL,
  gof_map = c("nobs", "r.squared"),
  coef_rename = c(
    "(Intercept)" = "Constant",
    "totwrk" = "Minutes Worked per Week"
  )
)
```

**Interpretation:**

A jobless person is predicted to sleep **3586** minutes per week \[**8.54** hours of daily sleep\].\
For every 100 minutes worked, 15 minutes of sleep are lost.

> (ii)

```{r}
#| label: c3_ii

sleep_fall <- coef(c3m)["totwrk"] * 120

cat("Estimated change in sleep:", round(sleep_fall, 2), "minutes")
```

| Model | Regression Equation | Interpretation of $\hat \beta_1$ | Name |
|------------------|------------------|-------------------|------------------|
| Level-Level | $y = \hat \beta_0 + \hat \beta_1 x$ | $\Delta \hat y = \hat \beta \Delta x$ | Marginal Effect |
| Log-Level | $\log(y) = \hat \beta_0 + \hat \beta_1 x$ | $\% \Delta \hat y \approx (100\hat \beta_1) \Delta x$ | Semi-elasticity |
| Level-Log | $y = \hat \beta_0 + \hat \beta_1 \log(x)$ | $\Delta \hat y \approx (\hat \beta_1)/100 \% \Delta x$ | \- |
| Log-Log | $\log(y) = \hat \beta_0 + \hat \beta_1 \log(x)$ | $\% \Delta \hat y \approx \hat \beta_1 \% \Delta x$ | Elasticity |

**Interpretation:**

Working an extra two hours per week will reduce weekly sleep by **18** minutes \[2.6 minutes per night\]. This is practically insignificant.

## C4 \[WAGE2\]

```{r}
#| label: c4_data

c4 <- wage2
kable(head(c4))
```

> (i)

```{r}
#| label: c4_i

c4 |>
  summarise(
    Avg_Salary = mean(wage),
    Avg_IQ = mean(IQ),
    SD_IQ = sd(IQ)
  ) |>
  round(2)
```

**Interpretation:**

The sample average wage ($\overline{wage}$) is **957.95**, with an average IQ ($\overline{IQ}$) of **101.28**. The sample standard deviation for IQ $\hat \sigma_{IQ}$ is **15.05**

> (ii)

```{r}
#| label: c4_ii

c4m <- lm(wage ~ IQ, data = c4)

modelsummary(
  c4m,
  fmt = 2,
  statistic = NULL,
  gof_map = c("nobs", "r.squared"),
  coef_rename = c(
    "(Intercept)" = "Constant"
  )
)
```

```{r}
#| label: c4_ii_2

IQ_inc <- coef(c4m)["IQ"] * 15

cat("Estimated change in wage:", round(IQ_inc, 2), "dollars")
```

**Interpretation:**

An increase in IQ by 15 is associated with an in increase in predicted wage by **\$124.55**. IQ explains only **9.6%** of the variation in wage ; the remaining variation is unexplained by the model.

> (iii)

```{r}
#| label: c4_iii

c4m2 <- lm(lwage ~ IQ, data = c4)

modelsummary(
  c4m2,
  fmt = 3,
  statistic = NULL,
  gof_map = c("nobs", "r.squared"),
  coef_rename = c(
    "(Intercept)" = "Constant"
  )
)
```

```{r}
#| label: c4_iii_2

IQ_inc <- coef(c4m2)["IQ"] * 15 * 100

cat("Estimated change in wage:", round(IQ_inc, 2), "percent")
```

**Interpretation:**

Each point increase is associated with an increase in wage by **0.9%**; 15 point increase in IQ will have a predicted wage increase by approximately **13%** percent.

## C5 \[RDCHEM\]

```{r}
#| label: c5_data

c5 <- rdchem
kable(head(c5))
```

> (i)

$$\log(rd) = \beta_0 + \beta_1 \log(sales) + u$$

**Interpretation:**

The parameter $\beta_1$ represents elasticity

> (ii)

```{r}
#| label: c5_i

c5m <- lm(lrd ~ lsales, data = c5)

modelsummary(
  c5m,
  fmt = 2,
  statistic = NULL,
  gof_map = c("nobs", "r.squared"),
  coef_rename = c(
    "(Intercept)" = "Constant",
    "lsales" = "Log Sales"
  )
)
```

**Interpretation:**

A **1%** change in sales is associated with a *1.08%* change in expenditures.

## C6 \[MEAP93\]

```{r}
#| label: c6_data

c6 <- meap93
kable(head(meap93))
```

> (i)

**Interpretation:**

I believe dollar has a diminishing effect on passing rate. Most students with limited source can pass, but scoring maximum mark takes very high resources. Also a constant effect may predict students can pass the maximum score.

> (ii)

**Interpretation:**

$$\begin{align}
\Delta math10 &= \beta_1 \cdot \Delta \log(expend) \\
\Delta math10 &\approx \beta_1 \cdot \left( \frac{\Delta expend}{expend} \right) \\
\Delta math10 &\approx \beta_1 \cdot (0.10) \\
\Delta math10 &\approx \frac{\beta_1}{10}
\end{align}$$

> (iii)

```{r}
#| label: c6_iii

c6m <- lm(math10 ~ lexpend, data = c6)

modelsummary(
  c6m,
  fmt = 2,
  statistic = NULL,
  gof_map = c("nobs", "r.squared"),
  coef_rename = c(
    "(Intercept)" = "Constant",
    "lexpend" = "Student Spending"
  )
)
```

> (iv)

**Interpretation:**

Estimated effect for spending effect is **11.16**. If spending increases by 10%, math pass rate increase by **1.116** percentage points.

> (v)

```{r}
#| label: c6_v

c6m |>
  augment() |>
  select(math10, .fitted) |>
  pivot_longer(everything()) |>
  ggplot(aes(x = value)) +
  geom_histogram(bins = 40) +
  facet_wrap(~name)
```

**Interpretation:**

We don't have to worry about passing maximum rate because predicted rates are between the interval **\[20, 30\]** while original rates had maximum rate below 80.

## C7 \[CHARITY\]

```{r}
#| label: c7_data

c7 <- charity
kable(head(c7))
```

> (i)

```{r}
#| label: c7_i

c7 |>
  summarise(
    Avg_Gift = mean(gift),
    No_Gift_Pct = mean(gift == 0) * 100
  ) |>
  round(2)
```

**Interpretation:**

Average gift in the sample is **7.44** with **60%** of the sample having no gifts.

> (ii)

```{r}
#| label: c7_ii

c7 |>
  summarise(
    Min_Mailing = min(mailsyear),
    Avg_Mailing = mean(mailsyear),
    Max_Mailing = max(mailsyear)
  ) |>
  round(2)
```

**Interpretation:**

Minimum number of mailings per year is **0.25**. The average is **2.05** and maximum is **3.05**

> (iii)

```{r}
#| label: c7_iii

c7m <- lm(gift ~ mailsyear, data = c7)

modelsummary(
  c7m,
  fmt = 2,
  statistic = NULL,
  gof_map = c("nobs", "r.squared"),
  coef_rename = c(
    "(Intercept)" = "Constant",
    "mailsyear" = "Mailings per Year"
  )
)
```

> (iv)

**Interpretation:** For each extra mailing per year, gift is predicted to increase by **2.65**. If each mailing costs 1 guilder, net gain on each mail is **2.65 - 1 = 1.65**. This does not mean that charity makes a net gain on every mailing, it means that, on average, net gain will be 1.65.

> (v)

```{r}
#| label: c7_v

c7m |>
  augment() |>
  summarise(
    Min_Prediction = min(.fitted)
  ) |>
  round(2)
```

**Interpretation:**

Minimum predicted gift is **2.68** coming from the observation with 0.25 mailings per year ($2.01 + 2.65(0.25) = 2.68$). Predicting a zero is impossible with this model because $\hat \beta_0$ and $\hat \beta_1$ are both greater than zero; least possible value is $\hat \beta_0 = 2.01$

## C8 \[Simulation\]

> (i)

```{r}
#| label: c8_i

set.seed(42)

x_i <- runif(500, min = 0, max = 10)

cat(round(mean(x_i), 2), round(sd(x_i), 2))
```

**Interpretation:**

The sample mean of $x_i$ is 4.9 and the sample standard deviation is 2.89. These values are close to, but not exactly equal to, the theoretical population mean ($E[x] = 5$) and population standard deviation ($\sigma_x \approx 2.93$) of a Uniform(0,10) distribution due to finite sampling variance.

> (ii)

```{r}
#| label: c8_ii

set.seed(42)

u_i <- rnorm(500, mean = 0, sd = 6)

cat(round(mean(u_i), 2), round(sd(u_i), 2))
```

**Interpretation:**

The sample average of $u_i$ is -0.18, which is not exactly zero. While the population expectation is exactly zero ($E[u] = 0$), a finite sample of 500 random draws will inherently contain sampling error, preventing the sample mean from perfectly matching the population mean. The sample standard deviation is 5.83, which closely approximates the population standard deviation of 6.

> (iii)

```{r}
#| label: c8_iii

y_i <- 1 + (2 * x_i) + u_i

c8m <- lm(y_i ~ x_i)

modelsummary(
  c8m,
  statistic = NULL,
  gof_map = list(),
  coef_rename = c(
    "(Intercept)" = "Constant",
    "x_i" = "x"
  )
)
```

**Interpretation**

$\hat{\beta}_0$ is estimated to be 0.901 and $\hat{\beta}_1$ is estimated to be 1.983. They differ from the true population parameters ($\beta_0 = 1, \beta_1 = 2$) due to sampling error. The intercept deviates because the sample mean of the errors ($\bar{u}$) is not exactly zero. More importantly, the slope deviates because the sample covariance between $x_i$ and $u_i$ is not exactly zero.

> (iv)

```{r}
#| label: c8_iv

c8m |>
  augment() |>
  summarise(
    Sum_Resid = sum(.resid),
    Sum_x_resid = sum(x_i * .resid)
  ) |>
  round(2)
```

**Interpretation**

The sum of the residuals ($\sum \hat{u}_i$) and the sum of the product of the regressors and residuals ($\sum x_i \hat{u}_i$) are both effectively zero. This confirms that the algebraic properties of OLS hold perfectly in the sample.

> (v)

```{r}
#| label: c8_v

c8m |>
  augment() |>
  summarise(
    Sum_Error = sum(u_i),
    Sum_x_Error = sum(x_i * u_i)
  ) |>
  round(2)
```

**Interpretation:**

When using the unobserved population errors ($u_i$) instead of the OLS residuals ($\hat{u}_i$), the sums are far from zero ($\sum u_i = -90.14$ and $\sum x_i u_i = -512.34$). This highlights the fundamental mechanical difference between errors and residuals: OLS residuals are mathematically forced by the regression algorithm to sum to zero and be uncorrelated with $x$, whereas population errors are unobserved random variables that will not perfectly satisfy these orthogonal conditions in a finite sample.

> (vi)

```{r}
#| label: c8_vi

set.seed(43)

x_i <- runif(500, min = 0, max = 10)
u_i <- rnorm(500, mean = 0, sd = 6)
y_i <- 1 + (2 * x_i) + u_i

c8m2 <- lm(y_i ~ x_i)

modelsummary(
  c8m2,
  statistic = NULL,
  gof_map = list(),
  coef_rename = c(
    "(Intercept)" = "Constant",
    "x_i" = "x"
  )
)
```

**Interpretation:**

With a new random sample, the estimates change to $\hat{\beta}_0 = 1.022$ and $\hat{\beta}_1 = 2.040$. These differ from the estimates in part (iii) because every new sample draws a unique set of $x_i$ and $u_i$ values. Consequently, the new sample mean of $u_i$ and the new sample covariance between $x_i$ and $u_i$ have changed. The OLS estimator processes this new random sample variation, resulting in different deviations from the true population parameters.

::: callout-caution
Formula for the slope $\hat \beta_1$:

$$
\hat \beta_1 = \beta_1 + \dfrac{Cov(x,u)}{Var(x)}
$$

To be equal to the parameter, covariance must be = zero

Formula for $\hat \beta_0$:

$$
\hat \beta_0 = \beta_0 + (\beta_1 - \hat \beta_1) \bar x + \bar u
$$

To be equal to the parameter, average of error term must be exactly zero, and $\hat \beta_1 = \beta_1$
:::

## C9 \[COUNTYMURDERS\]

```{r}
#| label: c9_data

c9 <- countymurders |>
  filter(year == 1996)

kable(head(c9))
```

> (i)

## C10 \[CATHOLIC\]

## C11 \[GPA1\]

## C12 \[LOANAPP\]
