---
title: "The Simple Regression Model"
format: html
---

## Computer Exercises

```{r}
#| label: 'setup'
#| include: false
library(tidyverse)
library(wooldridge)
library(modelsummary)
```

### C1 \[401K\]

```{r}
#| label: c1_data
df_401k <- wooldridge::k401k
knitr::kable(head(df_401k))

```

> (i) Find the average participation rate and the average match rate in the sample of plans

```{r}
#| label: c1_i
df_401k |>
  summarise(
    Avg_Part_Rt = mean(prate),
    Avg_Match_Rt = mean(mrate)
  ) |>
  round(2)
```

**Interpretation:**  
On average, **87.36%** of eligible employees participate in their company's 401(k) plan. Their company matches *\$0.73* for every dollar an employee contributes.

> estimate the simple regression equation $$\widehat{prate} = \hat \beta_0 + \hat \beta_1 mrate$$

```{r}
#| label: c1_ii
model_401k <- lm(prate ~ mrate, data = df_401k)

modelsummary(
  model_401k,
  fmt = 2,
  statistic = NULL,
  gof_map = c("nobs", "r.squared"),
  coef_rename = c("mrate" = "Match Rate ($)", "(Intercept)" = "Constant")
)
```

**Interpretation:**
$$
\beta_0 = 83.03 \quad \beta_1 = 5.86 \quad R^2 = 0.075 \quad n = 1534 \quad
$$

> (iii) Interpret the intercept in your equation. Interpret the coefficient on mrate

**Interpretation:**  
If the match rate is 0, predicted participation rate is **83.08%**.\
For ever \$1 increase in match rate, participation is predicted to increase by **5.86 percentage points**.

> (iv) Find the predicted prate when mrate = 3.5. Is this a reasonable prediction?

```{r}
#| label: c1_iv
new_data <- data.frame(mrate = 3.5)
pred_prate <- predict(model_401k, newdata = new_data)
round(pred_prate, 2)
```

**Interpretation:**  
Model predicts a participation rate of **103.59%** which exceeds maximum possible value of 100%

> (v) How much of the variation in prate is explained by mrate? Is this a lot in your opinion?

**Interpretation:**  
Match rate explains only about **7.5%** of the variation in participation, while the remaining 92.5% are due to other factors.

### C2 \[CEOSAL2\]

```{r}
#| label: c2_data
df_ceosal2 <- wooldridge::ceosal2
knitr::kable(head(df_ceosal2))
```

> (i) Find the average salary and the average tenure in the sample.

```{r}
#| label: c2_i
df_ceosal2 |>
  summarise(
    Avg_Salary = mean(salary),
    Avg_Tenure = mean(ceoten)
  ) |>
  round(2)
```

**Interpretation:**\
Average salary is **\$865,860** with an average tenure of **7.95** years.

> (ii) How many CEOs are in their first year as CEO? What is the longest tenure as a CEO?

```{r}
#| label: c2_ii
df_ceosal2 |>
  summarise(
    Fresh_CEO = sum(ceoten == 0),
    Veteran = max(ceoten)
  )
```

**Interpretation:**\    
There are **5** CEOs in their first year, and the longest serving CEO has been in charge for **37** years.

> (iii) Estimate the regression model $$ \log(salary) = \beta_0 + \beta_1 ceoten + u$$

```{r}
#| label: c2_iii
model_ceosal2 <- lm(lsalary ~ ceoten, data = df_ceosal2)

modelsummary(
  model_ceosal2,
  fmt = 2,
  statistic = NULL,
  gof_map = c("nobs", "r.squared"),
  coef_rename = c("ceoten" = "Tenure", "(Intercept)" = "Constant")
)
```

**Interpretation:**\
For every 1 additional year of tenure as CEO, salary is predicted to increase by **1%**

::: callout-note
Looks like CEO aren't driven by loyalty, neither should you
:::

### C3 \[SLEEP75\]

```{r}
#| label: c3_data
df_sleep75 <- wooldridge::sleep75
knitr::kable(head(df_sleep75))
```

> (i) Estimate the model and reprot results 

```{r}
#| label: c3_i
model_sleep75 <- lm(sleep ~ totwrk, data = df_sleep75)

modelsummary(
  model_sleep75,
  fmt = 2,
  statistic = NULL,
  gof_map = c("nobs", "r.squared"),
  coef_rename = c("totwrk" = "Minutes Worked per Week", "(Intercept)" = "Constant")
)
```

**Interpretation:**\  
A jobless person is predicted to sleep **3586** minutes per week [**8.54** hours of daily sleep].  
For every 100 minutes worked, 15 minutes of sleep are lost.

> (ii) If totwrk increases by 2 hours, by how much is sleep estimated to fall?

```{r}
#| label: c3_ii
new_data <- data.frame(totwrk = 120)
prediction <- predict(model_sleep75, newdata = new_data)

result <- model_sleep75$coefficients[1]
paste(round(prediction, 2), round(prediction - result, 2))
```
**Interpretation:**\  
Working an extra two hours per week will reduce weekly sleep by **18** minutes [2.6 minutes per night]. This is practically insignficant.
