---
title: "Multiple Regression Analysis: OLS Asymptotics"
format: html
---

## Computer Exercises

```{r}
#| label: 'setup'
#| include: false
library(tidyverse)
library(wooldridge)
library(modelsummary)
library(car)
library(broom)
```

### C1 \[WAGE1\]

```{r}
#| label: c1_data
df_wage1 <- wooldridge::wage1
knitr::kable(head(df_wage1))
```

> (i) Estimate the equation and plot the residuals $$wage = \beta_0 + \beta_1 educ + \beta_2 exper + \beta_3 tenure + u$$

```{r}
#| label: c1_i
model_wage1 <- lm(
  wage ~ educ + exper + tenure, data = df_wage1
)

ggplot( data = model_wage1, aes(x = .resid)) +
  geom_histogram(bins = 30)
```

**Interpretation:**

The residuals are highly right-skewed and clearly violate the assumption of normality.

> (ii) Use log wage as dependent variable

```{r}
#| label: c1_ii
model_lwage <- lm(
  lwage ~ educ + exper + tenure, data = df_wage1
)

ggplot( data = model_lwage, aes(x = .resid)) +
  geom_histogram(bins = 30)
```

**Interpretation:**

Residuals of the log wage model seem symmetric with few outliers.

> (iii) Would you say that Assumption MLR.6 is closer to being satisfied for the level-level model or the log-level model?

```{r}
#| label: c1_iii
shapiro_level <- shapiro.test(resid(model_wage1))
print(shapiro_level)

shapiro_log <- shapiro.test(resid(model_lwage))
print(shapiro_log)
```

**Interpretation:**

The log-level model is much closer to satisfying MLR.6. The level-level residuals are heavily right-skewed, blatantly violating the normality assumption. By taking the log of wages, the distribution of residuals becomes roughly symmetric and bell-shaped, which aligns much more closely with the normality requirement of MLR.6.

::: callout-tip
using `model$residuals` will not work with all models. Instead use `resid(model)`

To get coefficients use `tidy(model)`

To get Model-Level Statistics use `glance(model)`

To get Observation-Level Statistics use `augment(model)`

For presentation, keep using `modelsummary(model)`
:::

### C2 \[GPA2\]

```{r}
#| label: c2_data
df_gpa2 <- wooldridge::gpa2
knitr::kable(head(df_gpa2))
```

> (i) Estimate the equation $$colgpa = \beta_0 + \beta_1 hsperc + \beta_2 sat +u$$

```{r}
#| label: c2_i
model_gpa2 <- lm(
  colgpa ~ hsperc + sat, data = df_gpa2
)

modelsummary(
  model_gpa2,
  fmt = 2,
  shape = term ~ model + statistic,
  estimate = "{estimate}{stars}",
  statistic = c("std.error", "statistic", "p.value"),
  gof_map = c("nobs", "r.squared", "adj.r.squared"),
  coef_rename = c(
    "(Intercept)" = "Constant",
    "hsperc"    = "High School Percentage",
    "sat" = "Sat Score"
  )
)
```

> (ii) Reestimate the equation using first 2070 observations

```{r}
df_gpa2_sub <- df_gpa2 |> 
  slice(1:2070)

model_gpa2_sub <- lm(
  colgpa ~ hsperc + sat, data = df_gpa2_sub
)

modelsummary(
  model_gpa2_sub,
  fmt = 2,
  shape = term ~ model + statistic,
  estimate = "{estimate}{stars}",
  statistic = c("std.error", "statistic", "p.value"),
  gof_map = c("nobs", "r.squared", "adj.r.squared"),
  coef_rename = c(
    "(Intercept)" = "Constant",
    "hsperc"    = "High School Percentage",
    "sat" = "Sat Score"
  )
)
```

> \(iii\) Find the ratio of the standard errors on hsperc from parts (i) and (ii)

```{r}
#| label: c2_iii
se_full <- tidy(model_gpa2) |>
  filter(term == "hsperc") |>
  pull(std.error)

se_sub <- tidy(model_gpa2_sub) |>
  filter(term == "hsperc") |>
  pull(std.error)

print(se_sub / se_full)
```

**Interpretation:**

The ratio is approximately 1.3. This is close to the theoretical penalty of $\sqrt{2} \approx 1.414$. The penalty is not exactly 1.414 because while subsetting the data cut $SST_j$ roughly in half, it also caused slight fluctuations in the residual variance ($\hat{\sigma}$) and the variable's collinearity ($R^2_j$). $$SE(\hat \beta_j) = \frac{\hat \sigma}{\sqrt{SST_j(1-R^2_j)}}$$

### C3 \[BWGHT\]

```{r}
#| label: c3_data
df_bwght <- wooldridge::bwght
knitr::kable(head(df_bwght))
```

> compute the LM statistic for testing whether motheduc and fatheduc are jointly significant. $$\begin{align}bwght =& \beta_0 + \beta_1 cigs + \beta_2 parity + \beta_3 faminc \\&+ \beta_4 motheduc + \beta_5 fatheduc + u\end{align}$$

**The Math Way:**

To do Lagrange Multiplier Test:

1.  Filter the data from missing data
2.  Estimate the restricted model
3.  Extract the residuals $\tilde u$
4.  Regress $\tilde u$ on all independent variables
5.  Extract $R^2_u$ and $n$
6.  $LM = n \cdot R^2_u$
7.  $LM \sim \chi^2_q$ where q is number of restrictions

```{r}
#| label: c3_i
df_clean <- df_bwght |>
  drop_na(bwght, cigs, parity, faminc, motheduc, fatheduc)

model_r <- lm(bwght ~ cigs + parity + faminc,
              data = df_clean)

df_clean$u_tilde <- resid(model_r)

model_aux <- lm(
  u_tilde ~ cigs + parity + faminc + motheduc + fatheduc,
  data = df_clean)

n <- nobs(model_aux)
r2_aux <- summary(model_aux)$r.squared

lm_stat <- n * r2_aux
p_val <- pchisq(lm_stat, df = 2, lower.tail = FALSE)

cat("LM Statistic: ", lm_stat, "\n")
cat("P-value:      ", p_val, "\n")
```

**Interpretation:**

Calculated LM Statistic is 2.8 with p value 0.24. We fail to reject the null hypothesis; education is not jointly statistically significant.

::: callout-important
If we wanted to test if education = 0.5 not just zero, we will modify y from step two

`df_clean <- df_clean |>  mutate(bwght_tilde = bwght - (0.5 * motheduc) - (0.5 * fatheduc))`
:::
