---
title: "Multiple Regression Analysis: Inference"
format: html
---

## Computer Exercises

```{r}
#| label: 'setup'
#| include: false
library(tidyverse)
library(wooldridge)
library(modelsummary)
library(car)
```

### C1 \[VOTE1\]

```{r}
#| label: c1_data
df_vote1 <- wooldridge::vote1
knitr::kable(head(df_vote1))
```

> (i) To study whether campaign expenditures affect election outcomes: voteA is the percentage of the vote received by Candidate A, expendA and expendB are campaign expenditures by Candidates A and B, and prtystrA is a measure of party strength for Candidate A $$ voteA = \beta_0 + \beta_1 \log(expendA) + \beta_2 \log(expendB) + \beta_3 prtystrA + u$$ Interpret $\beta_1$

```{r}
#| label: c1_i
model_vote1 <- lm(voteA ~ lexpendA + lexpendB + prtystrA, data = df_vote1)

modelsummary(
  model_vote1,
  fmt = 2,
  shape = term ~ model + statistic,
  estimate = "{estimate}{stars}",
  statistic = c("std.error", "statistic", "p.value"),
  gof_map = c("nobs", "r.squared", "adj.r.squared"),
  coef_rename = c(
    "(Intercept)" = "Constant",
    "lexpendA"    = "Log(Expenditures A)",
    "lexpendB" = "Log(Expenditures B)",
    "prtystrA"    = "Party Strength A"
  )
)
```

**Interpretation:**

$\beta_1 = 7.23$, a 1% increase in candidate A's expenditures is associated with 0.07 percentage point increase in the share of the vote for candidate A, holding other factors fixed.

::: callout-tip
**The Two Times Rule**

If the absolute estimate is at least \*\*twice as large\*\* as its standard error, it is statistically significant
:::

> (ii) In terms of the parameters, state the null hypothesis that a 1% increase in A’s expenditures is offset by a 1% increase in B’s expenditures.

**Interpretation:**

$$\begin{align}
H_0: \beta_1 + \beta_2 &= 0\\
H_1: \beta_1 + \beta_2 &\neq 0
\end{align}$$

> (iii) Estimate the given model using the data in VOTE1 and report the results in usual form. Do A’s expenditures affect the outcome? What about B’s expenditures? Can you use these results to test the hypothesis in part (ii)?

**Interpretation:**

All the independent variables are statistically significant, indicating that $\beta \neq 0$. We still can't test the hypothesis from part 2 because we don't have the covariance needed

$$ 
se(\hat \beta_1 + \hat \beta_2) = \sqrt{Var(\hat \beta_1) + Var(\hat \beta_2) + 2 Cov(\hat \beta_1, \hat \beta_2)}
$$

> (iv) Estimate a model that directly gives the t statistic for testing the hypothesis in part (ii). What do you conclude? (Use a two-sided alternative.)

The easy way:

```{r}
#| label: c1_iv
joint_test <- linearHypothesis(model_vote1,
                               "lexpendA + lexpendB = 0")
print(joint_test)
```

**Interpretation:**

**p \> 0.05**, we fail to reject the hypothesis; if both candidates double their budget, election outcome stays the same.

The math way:

We define a new estimator $\theta = \beta_1 + \beta_2$ which means that $\beta_1 = \theta - \beta_2$ Substitute into the original equation: $$
\begin{align}
voteA &= \beta_0 + \beta_1 \log(expendA) + \beta_2 \log(expendB) + \beta_3 prtystrA + u \\
voteA &= \beta_0 + (\theta - \beta_2) \log(expendA) + \beta_2 \log(expendB) + \beta_3 prtystrA + u \\
voteA &= \beta_0 + \theta \log(expendA) - \beta_2 \log(expendA) + \beta_2 \log(expendB) + \beta_3 prtystrA + u \\
voteA &= \beta_0 + \theta \log(expendA) + \beta_2[\log(expendB) - \log(expendA)]   + \beta_3 prtystrA + u 
\end{align}$$

```{r}
#| label: c1_iv_2

model_vote1_reparameterized <- lm(
  voteA ~ lexpendA + I(lexpendB - lexpendA) + prtystrA, data = df_vote1)

modelsummary(
  model_vote1_reparameterized,
  fmt = 2,
  shape = term ~ model + statistic,
  estimate = "{estimate}{stars}",
  statistic = c("std.error", "statistic", "p.value"),
  gof_map = c("nobs", "r.squared", "adj.r.squared"),
  coef_rename = c(
    "(Intercept)"             = "Constant",
    "lexpendA"                = "Log(Expenditures A) [\u03b8]",
    "lexpendB"                = "Log(Expenditures B) [\u03b22]",
    "I(lexpendB - lexpendA)"  = "Log(Exp B) - Log(Exp A) [\u03b22]",
    "prtystrA"                = "Party Strength A [\u03b23]"
  )
)
```

**Interpretation:**

$\theta = -0.53$, with a p-value of 0.32. We fail to reject the null hypothesis. This implies that $\beta_1$ and $\beta_2$ perfectly offset each other; if both candidates double their budgets, the net effect on the election outcome is practically zero.

### C2 \[LAWSCH85\]

```{r}
#| label: c2_data
df_lawsch85 <- wooldridge::lawsch85
knitr::kable(head(df_lawsch85))
```

> (i) State and test the null hypothesis that the rank of law schools has no ceteris paribus effect on median starting salary. $$ \log(salary) = \beta_0 + \beta_1 LSAT + \beta_2 GPA + \beta_3 \log(libvol) + \beta_4 \log(cost) + \beta_5 rank + u $$

**Interpretation:**

$$\begin{align}H_0: \beta_5 &= 0\\
H_1: \beta_5 &\neq 0\end{align}$$

```{r}
#| label: c2_i
model_lawschl85 <- lm(
  lsalary ~ LSAT + GPA + llibvol + lcost + rank,
  data = df_lawsch85)

modelsummary(
  model_lawschl85,
  shape = term ~ model + statistic,
  fmt = 3,
  estimate = "{estimate}{stars}",
  statistic = c("std.error", "statistic", "p.value"),
  gof_map = c("nobs", "r.squared", "adj.r.squared"),
  coef_rename = c(
    "(Intercept)" = "Constant",
    "LSAT"        = "LSAT Score",
    "GPA"         = "Undergrad GPA",
    "llibvol"     = "Log(Library Volumes)",
    "lcost"       = "Log(Annual Cost)",
    "rank"        = "Law School Rank"
  )
)

 
```

**Interpretation:**

Law School Rank is estimated to have a coefficient of -0.003 with a very low p value indicating rejection of null hypothesis. Moving down one spot in the law school rankings (e.g., from 10th to 11th) is associated with an estimated 0.3% reduction in median starting salary, holding all else constant.

> \(ii\) Are features of the incoming class of students—namely, LSAT and GPA—individually or jointly significant for explaining salary?

```{r}
#| label: c2_ii
joint_test <- linearHypothesis(model_lawschl85,
                               c("LSAT = 0", "GPA = 0"))

print(joint_test)
```

**Interpretation:**

Individually, GPA is highly significant while LSAT is not (likely due to collinearity with school rank). However, the joint F-test yields a highly significant p-value, meaning we reject the null hypothesis; together, the incoming class characteristics are jointly significant.

> \(iii\) Test whether the size of the entering class (clsize) or the size of the faculty (faculty) needs to be added to this equation; carry out a single test

```{r}
#| label: c2_iii
model_unrestricted <- lm(
  lsalary ~ LSAT + GPA + llibvol + lcost + rank + clsize + faculty, 
  data = df_lawsch85
)

joint_test <- linearHypothesis(model_unrestricted, c("clsize = 0", "faculty = 0"))
print(joint_test)
```

**Interpretation:**

Restricted model shows no significant difference from unrestricted model (p = 0.39). We fail to reject the null hypothesis, class size and faculty provide no additional explanatory power

> \(iv\) What factors might influence the rank of the law school that are not included in the salary regression?

**Interpretation:**

Unobserved factors absorbed by the rank variable might include the school's geographic location, the strength of its alumni network, and the socioeconomic background of the student body (such as family income or father's occupation)

### C3 \[HPRICE1\]

```{r}
#| label: c3_data
df_hprice1 <- wooldridge::hprice1
knitr::kable(head(df_hprice1))
```

> \(i\) Estimate and obtain a confidence interval for the percentage change in price when a 150 square foot bedroom is added to a house
>
> $$
> \log(price) = \beta_0 + \beta_1 sqrft + \beta_2 bdrms + u
> $$

```{r}
#| label: c3_i
model_hprice1 <- lm(
  lprice ~ sqrft + bdrms, data = df_hprice1
)

modelsummary(
  model_hprice1,
  shape = term ~ model + statistic,
  fmt = 2,
  estimate = "{estimate}{stars}",
  statistic = c("std.error", "statistic", "p.value"),
  gof_map = c("nobs", "r.squared", "adj.r.squared"),
  coef_rename = c(
    "(Intercept)"  = "Constant",
    "sqrft"        = "Square Foot",
    "bdrms"        = "Bedrooms"
  )
)

```

**The Easy Way**

```{r}
#| label: c3_i_2
deltaMethod(model_hprice1, "150 * sqrft + bdrms")
```

**Interpretation:**

Adding a 150 square foot bedroom is associated with an estimated **8.58%** increase in price. We are 95% sure the true value lies in the interval **\[3.33%, 13.83%\]**

> \(ii\) write $\beta_2$ in terms of $\theta$ and $\beta_1$

**The Math Way:**

To get the confidence Interval for a 150 square foot bedroom, let $\theta = 150 \times \beta_1 + \beta_2$.

$$
\begin{align}
\beta_2 &= \theta - 150 \beta_1\\
\log(price) &= \beta_0 + \beta_1(sqrft) + (\theta - 150 \beta_1)bdrms + u \\
\log(price) &= \beta_0 + \beta_1 (sqrft - 150 \cdot bdrms) + \theta(bdrms) + u
\end{align}
$$

> (iii) Obtain standard error for $\hat \theta$ and construct a 95% confidence Interval

We run regression to obtain its standard error and confidence interval

```{r}
#| label: c3_iii
df_hprice1$sqrft_adj <- df_hprice1$sqrft - 150 * df_hprice1$bdrms

model_theta <- lm(I(100*lprice) ~ sqrft_adj + bdrms, data = df_hprice1)




cm <- c("bdrms" = "Effect of 150 sqft Bedroom (%)")

modelsummary(
  model_theta,
  coef_map = cm,
  fmt = 2,
  estimate = "{estimate} [{conf.low}, {conf.high}]",
  statistic = NULL,   
  title = "Estimated Impact of Adding a 150 sqft Bedroom",
  gof_omit = ".*"     
)

```

**Interpretation:**

We scaled dependent variable by 100 to get percentage values for the log-level model. The estimated effect is 8.58% with a confidence interval of \[3.26%, 13.90%\].

::: callout-caution
`deltMethod` relies on asymptotic theory, confidence intervals are calculated using Z distribution. Critical value = **1.96**

`lm` uses t distribution. With 85 degrees of freedom, critical value = **1.988**
:::
