---
title: "Multiple Regression Analysis: Estimation"
format: html
---

```{r}
#| label: 'setup'
#| message: false
#| warning: false

library(tidyverse)
library(wooldridge)
library(knitr)
library(modelsummary)
library(broom)
```

## C1 [BWGHT]

> A problem of interest to health officials (and others) is to determine the effects of smoking during pregnancy on infant health. One measure of infant health is birth weight; a birth weight that is too low can put an infant at risk for contracting various illnesses. Since factors other than cigarette smoking that affect birth weight are likely to be correlated with smoking, we should take those factors into account. For example, higher income generally results in access to better prenatal care, as well as better nutrition for the mother. An equation that recognizes this is $$bwght = \beta_0 + \beta_1 cigs + \beta_2 faminc + u$$

```{r}
#| label: c1_data

c1 <- bwght
kable(head(c1))
```

> (i) What is the most likely sign for $\beta_2$?

**Interpretation:**

$\beta_2$ is expected to be **positive**. Higher family income $\rightarrow$ better nutrition $\rightarrow$ higher birth weight

> (ii) Do you think `cigs` and `faminc` are likely to be correlated? Explain why the correlation might be positive or negative.

**Interpretation:**

Yes, **negative** correlation. smoking rates is higher among lower-income demographics. Omitting family income will cause Omitted Variable Bias (OVB).

> (iii) Now, estimate the equation with and without `faminc`, using the data in BWGHT. Report the results in equation form, including the sample size and R-squared. Discuss your results, focusing on whether adding `faminc` substantially changes the estimated effect of `cigs` on `bwght`.

```{r}
#| label: c1_iii

c1m <- lm(bwght ~ cigs, data = c1)
c1m2 <- lm(bwght ~ cigs + faminc, data = c1)

c1models <- list(
  "Without Income" = c1m,
  "With Income" = c1m2
)

modelsummary(
  c1models,
  fmt = 2,
  statistic = NULL,
  gof_map = c("nobs", "r.squared"),
  coef_rename = c(
    "(Intercept)" = "Constant",
    "cigs" = "Cigarettes/Day",
    "faminc" = "Family Income ($1000s)"
  )
)
```

**Interpretation:**

Including income reduces the smoking coefficient from approximately $-0.514$ to $-0.463$. This indicates that the simple regression suffered from omitted variable bias.$$ \text{Bias} = \beta_2 \times \tilde \delta_1$$

1\. $\beta_2$: Positive, rich people have heavier babies.

2\. $\tilde \delta_1$: Negative, poor people smoke more

3\. Effect: Positive $\times$ Negative = **Negative Bias**

4\. $\beta_1$ is negative, negative bias will increase it from -0.46 to -0.51

## C2 [HPRICE1]

> Use the data in HPRICE1 to estimate the model $$price = \beta_0 + \beta_1sqrft + \beta_2bdrms + u$$ where `price` is the house price measured in thousands of dollars.

```{r}
#| label: c2_data

c2 <- hprice1
kable(head(c2))
```

> (i) Write out the results in equation form.

```{r}
#| label: c2_i

c2m <- lm(price ~ sqrft + bdrms, data = c2)

modelsummary(
  c2m,
  fmt = 2,
  statistic = NULL,
  gof_map = c("nobs", "r.squared"),
  coef_rename = c(
    "(Intercept)" = "Constant",
    "sqrft"       = "Size (sq. ft.)",
    "bdrms"       = "Number of Bedrooms"
  )
)
```

> (ii) What is the estimated increase in price for a house with one more bedroom, holding square footage constant?

**Interpretation:**

Holding square footage constant, adding one more bedroom increases price by \$15,200. This estimate is not useful because adding small rooms without adding square footage will make it feel cramped.

> (iii) What is the estimated increase in price for a house with an additional bedroom that is 140 square feet in size? Compare this to your answer in part (ii).

```{r}
#| label: c2_iii

base_house <- data.frame(sqrft = 140, bdrms = 1)
upgraded_house <- data.frame(sqrft = 280, bdrms = 2)

price_1 <- predict(c2m, newdata = base_house)
price_2 <- predict(c2m, newdata = upgraded_house)

cat(round(price_2 - price_1, 2))
```

**Interpretation:**

The estimates increase is **\$33,180**. This represents the combined value of adding a room that is 140 square feet $$
140 \times 0.13 + 15.2 = 33.2$$

> (iv) What percentage of the variation in price is explained by square footage and number of bedrooms?

**Interpretation:**

**63.2%** of the variation in house prices are explained by the two variables.

> (v) The first house in the sample has `sqrft` = 2,438 and `bdrms` = 4. Find the predicted selling price

```{r}
#| label: c2_v

cat(c2m$fitted.values[1])
```

**Interpretation:**

Predicted selling price is **\$354,605**

> (vi) The actual selling price of the first house in the sample was \$300,000 Find the residual for this house. Does it suggest that the buyer underpaid or overpaid for the house?

```{r}
#| label: c2_vi

cat(c2m$residuals[1])
```

**Interpretation:**

Residual is **\$-54,605**. The buyer underpaid for the house.

## C3 [CEOSAL2]

> The file CEOSAL2 contains data on 177 chief executive officers and can be used to examine the effects of firm performance on CEO salary.

```{r}
#| label: c3_data

c3 <- ceosal2
kable(head(c3))
```

> (i) Estimate a model relating annual salary to firm sales and market value. Make the model of the constant elasticity variety for both independent variables.

```{r}
#| label: c3_i

c3m <- lm(lsalary ~ lsales + lmktval, data = c3)

modelsummary(
  c3m,
  fmt = 2,
  statistic = NULL,
  gof_map = c("nobs", "r.squared"),
  coef_rename = c(
    "(Intercept)" = "Constant",
    "lsales" = "Log(Sales)",
    "lmktval" = "Log(Market Value)"
  )
)
```

> (ii) Add `profits` to the model from part (i). Why can this variable not be included in logarithmic form? Would you say that these firm performance variables explain most of the variation in CEO salaries?

```{r}
#| label: c3_ii

ggplot(c3, aes(x = profits)) +
  geom_histogram(bins = 30)
```

```{r}
#| label: c3_ii2

c3m2 <- lm(lsalary ~ lsales + lmktval + profits, data = c3)

modelsummary(
  c3m2,
  fmt = 2,
  statistic = NULL,
  gof_map = c("nobs", "r.squared"),
  coef_rename = c(
    "(Intercept)" = "Constant",
    "lsales" = "Log(Sales)",
    "lmktval" = "Log(Market Value)",
    "profits" = "Profits"
  )
)
```

**Interpretation:**

Profits can't be added in log form because it has non positive values. The model explains **30%** of the variation in the salaries. The remaining 70% is determined by other factors like tenure, unobserved CEO talent, or corporate politics.

> (iii) Add the variable `ceoten` to the model in part (ii). What is the estimated percentage return for another year of CEO tenure, holding other factors fixed?

```{r}
#| label: c3_iii

c3m3 <- lm(
  lsalary ~ lsales + lmktval + profits + ceoten,
  data = c3
)

modelsummary(
  c3m3,
  fmt = 2,
  statistic = NULL,
  gof_map = c("nobs", "r.squared"),
  coef_rename = c(
    "(Intercept)" = "Constant",
    "lsales" = "Log(Sales)",
    "lmktval" = "Log(Market Value)",
    "profits" = "Profits",
    "ceoten" = "Tenure"
  )
)
```

**Interpretation:**

$\beta_4$ = 0.01, one additional year of tenure is associated with **1%** increase in salary, holding firm performance constant.

> (iv) Find the sample correlation coefficient between the variables `log(mktval)` and `profits`. Are these variables highly correlated? What does this say about the OLS estimators?

```{r}
#| label: c3_iv

c3 |>
  summarise(
    Correlation = cor(lmktval, profits)
  ) |>
  round(2)
```

**Interpretation:**

Correlation = **0.78**. This high correlation between market value and profits indicates multicollinearity. While this does not bias our estimates or inflate the overall $R^2$, it makes it difficult to isolate the individual partial effects of profits versus market value.

## C4 \[ATTEND\]

> Use the data in ATTEND for this exercise.

```{r}
#| label: c4_data

c4 <- attend
kable(head(c4))
```

> (i) Obtain the minimum, maximum, and average values for the variables `atndrte`, `priGPA`, and `ACT`.

```{r}
#| label: c4_i

c4 |>
  summarise(
    across(
      c(atndrte, priGPA, ACT),
      list(
        min = \(x) min(x, na.rm = TRUE),
        max = \(x) max(x, na.rm = TRUE),
        avg = \(x) mean(x, na.rm = TRUE)
      )
    )
  ) |>
  round(2)
```

**Interpretation:**

Attendance rate lies in the range \[6.25, 100\] with an average of 81.71. Prior GPA is in the interval \[0.86, 3.93\] with an average of 2.59. ACT score is in the interval \[13, 32\] with an average of 22.51.

> (ii) Estimate the model $$atndrte = \beta_0 + \beta_1 priGPA + \beta_2 ACT + u$$ and write the results in equation form. Interpret the intercept. Does it have a useful meaning?

```{r}
#| label: c4_ii

c4m <- lm(atndrte ~ priGPA + ACT, data = c4)

modelsummary(
  c4m,
  fmt = 2,
  statistic = NULL,
  gof_map = c("nobs", "r.squared"),
  coef_map = c(
    "(Intercept)" = "Constant",
    "priGPA"      = "Prior GPA",
    "ACT"         = "ACT Score"
  )
)
```

```{r}
#| label: c4_ii_2

c4 |>
  select(atndrte, priGPA, ACT) |>
  pivot_longer(cols = c(priGPA, ACT), names_to = "variable", values_to = "value") |>
  ggplot(aes(x = value, fill = atndrte > 80)) +
  geom_histogram(bins = 20, position = "identity", alpha = 0.6) +
  facet_wrap(~variable, scales = "free") +
  theme_minimal() +
  scale_x_continuous(n.breaks = 10) +
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust = 1))
```

**Interpretation:**

$\hat \beta_0 = 75.7$. A student with zero ACT score and zero prior GPA is expected to have attendance rate = **75.7**. The intercept has no useful meaning because regression is extrapolating.

> (iii) Discuss the estimated slope coefficients. Are there any surprises?

**Interpretation**

ACT score has a slope of **-1.72**. This is surprising as a higher score is usually correlated with higher attendance.

> (iv) What is the predicted atndrte if `priGPA` = 3.65. and `ACT` = 20? What do you make of this result? Are there any students in the sample with these values of the explanatory variables?

```{r}
#| label: c4_iv

c4 |>
  filter(ACT == 20 & priGPA == 3.65) |>
  summarise(
    Count = n()
  )
```

```{r}
#| label: c4_iv_2

df <- data.frame(priGPA = 3.65, ACT = 20)

prediction <- predict(c4m, newdata = df)
cat(prediction)
```

**Interpretation:**

A student with prior GPA of 3.65 and ACT score of 20 is predicted to have attendance rate of **104**; which passes maximum value of 100. The model has no students with such values so it is extrapolating.

> (v) If Student A has `priGPA` = 3 1. and `ACT` = 21 and Student B has `priGPA = 2.1`. and `ACT = 26`, what is the predicted difference in their attendance rates?

```{r}
#| label: c4_v

dfA <- data.frame(priGPA = 3.1, ACT = 21)
dfB <- data.frame(priGPA = 2.1, ACT = 26)

predictionA <- predict(c4m, newdata = dfA)
predictionB <- predict(c4m, newdata = dfB)

cat(predictionA - predictionB)
```

**Interpretation:**

The expected difference in attendance rate between the two students is **25.8**.

## C5 \[WAGE1\]

> (i) Confirm the partialling out interpretation of the OLS estimates by explicitly doing the partialling out for Example 3.2. $$\widehat{\log(wage)} = 0.284 + 0.092 educ + 0.0041 exper + 0.022 tenure$$ This first requires regressing `educ` on `exper` and `tenure` and saving the residuals, $\hat u$ Then, regress `log(wage)` on $\hat u$. Compare the coefficient on $\hat u$ with the coefficient on educ in the regression of log(wage) on educ, exper, and tenure.

```{r}
#| label: c5_data

c5 <- wage1
kable(head(c5))
```

```{r}
#| label: c5_i

c5m <- lm(educ ~ exper + tenure, data = c5)

c5 <- c5 |>
  mutate(
    u_hat = resid(c5m)
  )

c5m2 <- lm(lwage ~ u_hat, data = c5)

modelsummary(
  c5m2,
  fmt = 2,
  statistic = NULL,
  gof_map = c("nobs", "r.squared"),
  coef_map = c(
    "(Intercept)" = "Constant",
    "u_hat" = "Residuals"
  )
)
```

**Interpretation:**

Slope of the residual term is same as that of education **(0.09)**. This proves that multiple regression computes partial effect of each independent variable.

::: callout-note
### Frisch-Waugh-Lovell Theorem

In a multiple regression, the coefficient for $x_1$ represents partial effect of $x_1$ on $y$ after removing overlop with other variables:

1.  Regress: $x_1 = \alpha_0 + \alpha_1 x_2 + \alpha_2 x_3 + u$
2.  Extract: $\hat{u}$ (This is "$x_1$" that is completely uncorrelated with $x_2$ and $x_3$).
3.  Result: The slope $\beta_1$ from $y = \beta_0 + \beta_1 \hat{u} + e$ is identical to the coefficient in the full multiple regression.
4.  The Formula$$\hat{\beta}_{x_1} = \frac{\sum \hat{u}_i y_i}{\sum \hat{u}_i^2}$$
:::

## C6 \[WAGE2\]

> Use the data set in WAGE2 for this problem. As usual, be sure all of the following regressions contain an intercept.

```{r}
#| label: c6_data

c6 <- wage2
kable(head(c6))
```

> (i) Run a simple regression of `IQ` on `educ` to obtain the slope coefficient, say, $\tilde \delta_1$

```{r}
#| label: c6_i

c6m <- lm(IQ ~ educ, data = c6)

modelsummary(
  c6m,
  fmt = 3,
  statistic = NULL,
  gof_map = c("nobs", "r.squared"),
  coef_map = c(
    "(Intercept)" = "Constant",
    "educ" = "Education"
  )
)
```

**Interpretation:**

The slope coefficient $\tilde \delta_1$ is 3.534.

> (ii) Run the simple regression of `log(wage)` on `educ` and obtain the slope coefficient $\tilde \beta_1$

```{r}
#| label: c6_ii

c6m2 <- lm(lwage ~ educ, data = c6)

modelsummary(
  c6m2,
  fmt = 3,
  statistic = NULL,
  gof_map = c("nobs", "r.squared"),
  coef_map = c(
    "(Intercept)" = "Constant",
    "educ" = "Education"
  )
)
```

**Interpretation:**

The slope coefficient $\tilde \beta_1$ is 0.060.

> (iii) Run the multiple regression of `log(wage)` on `educ` and `IQ`, and obtain the slope coefficients, $\hat \beta_1, \hat \beta_2$ respectively.

```{r}
#| label: c6_iii

c6m3 <- lm(lwage ~ educ + IQ, data = c6)

modelsummary(
  c6m3,
  fmt = 3,
  statistic = NULL,
  gof_map = c("nobs", "r.squared"),
  coef_map = c(
    "(Intercept)" = "Constant",
    "educ" = "Education",
    "IQ" = "IQ"
  )
)
```

**Interpretation:**

$\hat \beta_1$ is 0.039, while $\hat \beta_2$ is 0.006

> (iv) Verify that $\tilde \beta_1 = \hat \beta_1 + \hat \beta_2 \tilde \delta_1$

$$\begin{align*}
\tilde \beta_1 &= \hat \beta_1 + \hat \beta_2 \tilde \delta_1\\
&= 0.039 + 0.006 * 3.534\\
&= 0.06
\end{align*}$$

**Interpretation:** 

From simple regression, we had effect of education on wage to be $\tilde \beta_1$ = 0.06. This estimate suffers from omitted variable bias and can be calculated as: education partial effect $\hat \beta\_1 $ = 0.039, IQ effect on wage $\hat \beta\_2$ = 0.006, how much IQ varies with education $\tilde \delta_1$ = 3.534

::: callout-important

### Omitted Variable Decomposition

$$
\tilde \beta_1 = \hat \beta_1 + \hat \beta_2 \tilde \delta_1
$$

Where:

-   $\tilde \beta_1$ is the simple regression estimate for independent variable

-   $\tilde \delta_1$ is the simple regression estimate for omitted variable on independent variable

-   $\hat \beta_1$ is the partial effect estimate from multiple regression for independent variable

-   $\hat \beta_2$ is the omitted variable effect from multiple regression on dependent variable
:::

## C7 [MEAP93]

> Use the data in MEAP93 to answer this question.

```{r}
#| label: c7_data

c7 <- meap93
kable(head(c7))
```

> (i) Estimate the model $$math10 = \beta_0 + \beta_1 \log(expend) + \beta_2 lnchprg + u$$ and report the results in the usual form, including the sample size and R-squared. Are the signs of the slope coefficients what you expected? Explain.

```{r}
#| label: c7_i

c7m <- lm(math10 ~ lexpend + lnchprg, data = c7)

modelsummary(
  c7m,
  fmt = 3,
  statistic = NULL,
  gof_map = c("nobs", "r.squared"),
  coef_map = c(
    "(Intercept)" = "Constant",
    "lexpend" = "Log(Spending per Pupil)",
    "lnchprg" = "% Eligible for Free Lunch"
  )
)
```

**Interpretation:**

Free lunch program has a negative sign. It may be acting as a proxy variable for poverty. The constant has a negative score which is also below the minimum range of 0.

> (ii) What do you make of the intercept you estimated in part (i)? In particular, does it make sense to set the two explanatory variables to zero? [Hint: Recall that log(1) 0 = 0.]

The intercept is predicting the math score will be negative if students spend exactly one dollar and no one is eligible for lunch programs. The regression is extrapolating.

> (iii)  Now run the simple regression of `math10` on `log(expend)`, and compare the slope coefficient with the estimate obtained in part (i). Is the estimated spending effect now larger or smaller than in part (i)?

```{r}
#| label: c7_iii

c7m2 <- lm(math10 ~ lexpend, data = c7)

modelsummary(
  c7m2,
  fmt = 3,
  statistic = NULL,
  gof_map = c("nobs", "r.squared"),
  coef_map = c(
    "(Intercept)" = "Constant",
    "lexpend" = "Log(Spending per Pupil)"
  )
)
```

**Interpretation:**

Estimated spending now is 11.16 after it was 6.23.

> (iv) Find the correlation between `lexpend` = log(expend) and `lnchprg`. Does its sign make sense to you?

```{r}
#| label: c7_iv

cor(c7$lexpend, c7$lnchprg)
```

**Interpretation:**

Student spending and eligibility for lunch program are negatively correlated (-0.2). It makes sense if lunch program is a proxy for poverty.

> (v) Use part (iv) to explain your findings in part (iii).

**Interpretation:**

$$Bias = \beta_2 \times Correlation(X_1, X_2)$$

Estimate of lunch program on math score $\hat \beta_2$ is negative. And the correlation between  the two independent variables is negative; resulting in a positive bias that is added on $\hat \beta_1$

## C8 [DISCRIM]

> Use the data in DISCRIM to answer this question. These are zip code–level data on prices for various items at fast-food restaurants, along with characteristics of the zip code population, in New Jersey and Pennsylvania. The idea is to see whether fast-food restaurants charge higher prices in areas with a larger concentration of blacks.

> (i) Find the average values of `prpblck` and `income` in the sample, along with their standard deviations. What are the units of measurement of `prpblck` and `income`?

> (ii) Consider a model to explain the price of soda, `psoda`, in terms of the proportion of the population that is black and median income: $$psoda = \beta_0 + \beta_1 prpblck + \beta_2 income + u$$ Estimate this model by OLS and report the results in equation form, including the sample size and R-squared. (Do not use scientific notation when reporting the estimates.) Interpret the coefficient on `prpblck`. Do you think it is economically large?

> (iii) Compare the estimate from part (ii) with the simple regression estimate from `psoda` on `prpblck`. Is the discrimination effect larger or smaller when you control for income?

> (iv) A model with a constant price elasticity with respect to income may be more appropriate. Report estimates of the model $$\log(psoda) = \beta_0 + \beta_1 prpblck + \beta_2 \log(income) + u$$ If `prpblck` increases by 0.20 (20 percentage points), what is the estimated percentage change in `psoda`? (Hint: The answer is 2.xx, where you fill in the “xx.”)

> (v) Now add the variable `prppov` to the regression in part (iv). What happens to $\hat \beta_{prpblck}$

> (vi) Find the correlation between log(`income`) and `prppov`. Is it roughly what you expected?

> (vii) Evaluate the following statement: “Because log(`income`) and `prppov` are so highly correlated, they have no business being in the same regression.”

## C9 [CHARITY]

> Use the data in CHARITY to answer the following questions:

> (i) Estimate the equation $$gift + \beta_0 + \beta_1 mailsyear + \beta_2 giftlast + \beta_3 propresp + u$$ by OLS and report the results in the usual way, including the sample size and R-squared. How does the R-squared compare with that from the simple regression that omits `giftlast` and `propresp`?

> (ii) Interpret the coefficient on `mailsyear`. Is it bigger or smaller than the corresponding simple regression coefficient?

> (iii) Interpret the coefficient on `propresp`. Be careful to notice the units of measurement of `propresp`.

> (iv) Now add the variable `avggift` to the equation. What happens to the estimated effect of `mailsyear`?

> (v) In the equation from part (iv), what has happened to the coefficient on `giftlast`? What do you think is happening?

> (vi) In order to guarantee that one gets an unbiased estimator of the causal effect of mailings, what would you recommend to the directors of the charity? Assume that the charity is willing to take some time, and that there is a budget for sending out mailings.

## C10 [HTV]

> Use the data in HTV to answer this question. The data set includes information on wages, education, parents’ education, and several other variables for 1,230 working men in 1991.

> (i) What is the range of the `educ` variable in the sample? What percentage of men completed twelfth grade but no higher grade? Do the men or their parents have, on average, higher levels of education?

> (ii) Estimate the regression model $$educ = \beta_1 motheduc + \beta_2 fatheduc + u$$ by OLS and report the results in the usual form. How much sample variation in `educ` is explained by parents’ education? Interpret the coefficient on `motheduc`.

> (iii) Add the variable `abil` (a measure of cognitive ability) to the regression from part (ii), and report the results in equation form. Does “ability” help to explain variations in education, even after controlling for parents’ education? Explain.

> (iv) (Requires calculus) Now estimate an equation where `abil` appears in quadratic form: $$educ = \beta_0 + \beta_1 motheduc + \beta_2 fatheduc + \beta_3 abil + \beta_4 abil^2 + u$$ Using the estimated $\hat \beta_3, \hat \beta_4$, use calculus to find the value of `abil`, call it `abil*`, where educ is minimized. (The other coefficients and values of parents’ education variables have no effect; we are holding parents’ education fixed.) Notice that `abil` is measured so that negative values are permissible. You might also verify that the second derivative is positive so that you do indeed have a minimum.

> (v) Argue that only a small fraction of men in the sample have “ability” less than the value calculated in part (iv). Why is this important?

> (vi) If you have access to a statistical program that includes graphing capabilities, use the estimates in part (iv) to graph the relationship between the predicted education and `abil`. Set `motheduc` and `fatheduc` at their average values in the sample, 12.18 and 12.45, respectively.

## C11 [MEAPSINGLE]

> Use the data in MEAPSINGLE to study the effects of single-parent households on student math performance. These data are for a subset of schools in southeast Michigan for the year 2000. The socioeconomic variables are obtained at the ZIP code level (where ZIP code is assigned to schools based on their mailing addresses).

> (i) Run the simple regression of `math4` on `pctsgle` and report the results in the usual format. Interpret the slope coefficient. Does the effect of single parenthood seem large or small?

> (ii) Add the variables `lmedinc` and `free` to the equation. What happens to the coefficient on `pctsgle`? Explain what is happening.

> (iii) Find the sample correlation between `lmedinc` and `free`. Does it have the sign you expect?

> (iv) Does the substantial correlation between `lmedinc` and `free` mean that you should drop one from the regression to better estimate the causal effect of single parenthood on student performance? Explain.

> (v) Find the variance inflation factors (VIFs) for each of the explanatory variables appearing in the regression in part (ii). Which variable has the largest VIF? Does this knowledge affect the model you would use to study the causal effect of single parenthood on math performance?

## C12 [ECONMATH]

> The data in ECONMATH contain grade point averages and standardized test scores, along with performance in an introductory economics course, for students at a large public university. The variable to be explained is `score`, the final score in the course measured as a percentage.

> (i) How many students received a perfect score for the course? What was the average score? Find the means and standard deviations of `actmth` and `acteng`, and discuss how they compare.

> (ii) Estimate a linear equation relating `score` to `colgpa`, `actmth`, and `acteng`, where `colgpa` is measured at the beginning of the term. Report the results in the usual form.

> (iii) Would you say the math or English ACT score is a better predictor of performance in the economics course? Explain.

> (iv) Discuss the size of the R-squared in the regression.

## C13 [GPA1]

> Use the data in GPA1 to answer this question. We can compare multiple regression estimates, where we control for student achievement and background variables, and compare our findings with the difference-in-means estimate in Computer Exercise C11 in Chapter 2.

> (i) In the simple regression equation $$colGPA = \beta_0 + \beta_1 PC + u$$

> (ii) Now add the controls `hsGPA` and `ACT`—that is, run the regression `colGPA` on `PC`, `hsGPA`, and `ACT`. Does the coefficient on PC change much from part (ii)? Does $\hat \beta_{colGPA}$ make sense?

> (iii) In the estimation from part (ii), what is worth more: Owning a PC or having 10 more points on the ACT score?

> (iv) Now to the regression in part (ii) add the two binary indicators for the parents being college graduates. Does the estimate of $\beta_1$ change much from part (ii)? How much variation are you explaining in `colGPA`?

> (v) Suppose someone looking at your regression from part (iv) says to you, “The variables `hsGPA` and `ACT` are probably pretty highly correlated, so you should drop one of them from the regression.” How would you respond?

> (vi) For curiosity’s sake, compute the VIFs for hsGPA and ACT. Of what value is this information?

## C14 [APPLE]

> Use the data in APPLE to answer the following questions. These come from a survey where respondents were presented with the characteristics of an “ecologically-friendly” apple. They were also randomly presented with a price pair: the price per lb. of the new apple, `ecoprc`, and the price per lb. of regular apples, `regprc`. Each respondent was asked, at the given pair of prices, how many pounds of eco-friendly apples (`ecolbs`) would they buy (and how many pounds of regular apples).

> (i) How many respondents are there in the sample? How many report `ecolbs` of zero? What is the second most common value reported for `ecolbs`?

> (ii) Consider a demand equation for the eco-friendly apples: $$ecolbs = \beta_0 + \beta_1 ecoprc + \beta_2 regprc + u $$. What does economic theory predict for the signs of $\beta_1, \beta_2$?

> (iii) Estimate the equation by OLS and report the results in equation form. Are any of the fitted values negative? Does the estimated intercept answer an interesting question? Explain. What is the R-squared?

> (iv) If `ecoprc` increases by 0.10 (10 cents per lb.), what is the estimated change in `ecolbs`, holding `regprc` fixed? Do you think this is a practically large effect?

> (v) If `ecoprc` and `regprc` both increase by 0.10, argue that the estimated change in `ecolbs` is close to zero.

> (vi) Suppose you present your estimated equation to a colleague who says, “I’m concerned about the low R-squared; it probably means the OLS estimators are biased.” How would you respond?

> (vii) Confirm that `ecoprc` and `regprc` are strongly positively correlated. What is the correlation? Does this mean that you should exclude `regprc` from the regression in order to estimate the own price effect? What happens to your estimate of $\beta_1$ if you drop `regprc` and use simple regression?
