---
title: "The Nature of Econometrics and Economic Data"
format: html
---

```{r}
#| label: 'setup'
#| message: false
#| warning: false

library(tidyverse)
library(wooldridge)
library(knitr)
library(janitor)
```

## C1 \[Wage 1\]

```{r}
#| label: c1_data

c1 <- wage1
kable(head(c1))
```

> (i)

```{r}
#| label: c1_i

c1 |>
  summarise(
    Avg_Educ = mean(educ),
    Min_Educ = min(educ),
    Max_Educ = max(educ)
  ) |>
  round(2)
```

**Interpretation:**

The average education in this sample is approximately 12.5 years. The range is broad ranging from 0 years (no formal education) to 18 years of education (graduate level).

> (ii)

```{r}
#| label: c1_ii

c1 |>
  summarise(
    Avg_Wage = mean(wage, na.rm = TRUE)
  ) |>
  round(2)
```

**Interpretation:**

The average nominal wage is \$5.90. On its own, in 1976 dollars, it might seem low compared to modern standards, but we need to account for inflation to truly judge its purchasing power.

> (iii)

**Interpretation:**

CPI for 1976 is **\$56.9** and became **\$292.7** in 2022.

> (iv)

$$
\begin{align*}
{Wage}_{2022} &= {Wage}_{1976} \times \dfrac{CPI_{2022}}{CPI_{1976}}\\
&= 5.9 \times \dfrac{292.7}{56.9}\\
&= 30.35
\end{align*}
$$

**Interpretation:**

The average wage in this sample is equivalent to \$30.35 in 2022 dollars.The nominal 1976 average wage of \$5.90 is more than 2.5 times the 1976 minimum wage of \$2.30. Because the sample's average earnings vastly exceed the baseline wage of its own era, it is highly probable that this sample includes a significant proportion of skilled or experienced labor.

> (v)

```{r}
#| label: c1_v

c1 |>
  summarise(
    Count = n(),
    Avg_Wage = mean(wage) |> round(2),
    .by = female
  ) |>
  mutate(Gender = ifelse(female == 1, "Female", "Male")) |>
  select(Gender, Count, Avg_Wage)
```

**Interpretation:**

Although sample is well balanced with 252 females and 274 males, there is a notable wage gap between the genders.

## C2 \[BWGHT\]

```{r}
#| label: c2_data

c2 <- bwght
kable(head(c2))
```

> (i)

```{r}
#| label: c2_i

c2 |>
  summarise(
    Count = n(),
    Smokers = sum(cigs > 0),
    Non_Smokers = sum(cigs == 0)
  )
```

**Interpretation:**

The sample contains **1,388 women**. Out of these, **212 women** (approximately 15%) reported smoking at least one cigarette per day during pregnancy, while **1,176** reported not smoking at all.

> (ii)

```{r}
#| label: c2_ii

c2 |>
  summarise(
    Avg_Cigs_All = mean(cigs)
  ) |>
  round(2)
```

```{r}
#| label: c2_ii_2

ggplot(c2, aes(x = cigs)) +
  geom_bar()
```

**Interpretation:**

The sample average is **2.09** cigarettes per day. However, this is not a good measure of a "typical" woman because the vast majority of pregnant females in the sample (about 85%) are non-smokers. The median is 0.

> (iii)

```{r}
#| label: c2_iii

c2 |>
  summarise(
    Avg_Cigs_Smokers = mean(cigs[cigs > 0])
  ) |>
  round(2)
```

**Interpretation:**

A pregnant female who smokes consumes an average of **13.67** cigarettes per day. This is significantly higher than the sample average of 2.09 because this calculation excludes the overwhelming majority of women who bring the overall average down to near zero.

> (iv)

```{r}
#| label: c2_iv

c2 |>
  summarise(
    Avg_Fath_Educ = mean(fatheduc, na.rm = TRUE),
    Missing_Obs = sum(is.na(fatheduc))
  ) |>
  round(2)
```

**Interpretation:**

An average father has **13.19** years of education in the sample, with *196* missing data.

> (v)

```{r}
#| label: c2_v

c2 |>
  summarise(
    Avg_Fam_Inc = mean(faminc, na.rm = TRUE),
    SD_Fam_Inc = sd(faminc, na.rm = TRUE)
  ) |>
  round(2)
```

**Interpretation:**

Families included in the sample have an average income of **\$29,000** with a standard deviation of **\$18,700**

## C3 \[MEAP01\]

```{r}
#| label: c3_data

c3 <- meap01
kable(head(c3))
```

> (i)

```{r}
#| label: c3_i

c3 |>
  summarise(
    Max_math4 = max(math4),
    Min_math4 = min(math4)
  ) |>
  round(2)
```

**Interpretation:**

The sample includes math scores ranging from **0** to **100** indicating wide variety.

> (ii)

```{r}
#| label: c3_ii

c3 |>
  summarise(
    Total_Schools = n(),
    Perfect_Math_Count = sum(math4 == 100, na.rm = TRUE),
    Percentage = (Perfect_Math_Count / Total_Schools) * 100
  ) |>
  round(2)
```

**Interpretation:**

Out of the 1,823 schools in the sample, 38 schools (approximately 2.08%) achieved a perfect 100% pass rate on the fourth-grade math test.

> (iii)

```{r}
#| label: c3_iii

c3 |>
  summarise(
    Schools_at_50 = sum(math4 == 50, na.rm = TRUE)
  )
```

**Interpretation:**

There are exactly 17 schools with a pass rate of exactly 50%

> (iv)

```{r}
#| label: c3_iv

c3 |>
  drop_na(math4, read4) |>
  summarise(
    Avg_Math = mean(math4),
    Avg_Read = mean(read4)
  ) |>
  round(2)
```

**Interpretation:**

The average pass rate for math **71.90%** is significantly higher than for reading **60.06%**. This shows that the reading exam was harder in the sample.

> (v)

```{r}
#| label: c3_v

c3 |>
  drop_na(math4, read4) |>
  summarise(
    Correlation = cor(math4, read4)
  ) |>
  round(2)
```

**Interpretation:**

The correlation between the two exams is high **0.84** showing that performance in the two exams is similar.

> (vi)

```{r}
#| label: c3_vi

c3 |>
  drop_na(math4, read4) |>
  summarise(
    Mean_Expnd = mean(exppp),
    Std_Expend = sd(exppp)
  ) |>
  round(2)
```

**Interpretation:**

The average expenditure per pupil is **\$5,194.87**, with a standard deviation of **\$1,091.89** representing a significant spread (**21%** of the mean).

::: callout-note
### Coefficient of Variation (CV)

To know if the spread is significant or not, calculate coefficient of variation (cv). $$
CV = \dfrac \sigma \mu$$

Small (cv \< 10%)

Medium (cv 10% - 30%)

Large(cv \> 30%)
:::

> (vii)

```{r}
#| label: c3_vii

school_a <- 6000
school_b <- 5500


pct_diff <- ((school_a - school_b) / school_b) * 100
log_diff <- 100 * (log(school_a) - log(school_b))


results <- c(Exact_Pct = pct_diff, Log_Approx = log_diff)
results |> round(2)
```

**Interpretation:**

School A spends **9.09%** more per student than School B. The logarithmic approximation ($100 \times \Delta \log$) yields **8.70**, which is a reasonably close estimate.

::: callout-important
### Logarithmic Approximation for Percentage Differences

When comparing two cross-sectional observations (such as expenditures between two different schools), we calculate a **percentage difference**. The exact percentage difference of $x_1$ relative to $x_0$ is:

$$\% \text{Difference} = \frac{x_1 - x_0}{x_0} \times 100$$

We can approximate this difference using natural logarithms:

$$\% \text{Difference} \approx 100 \times [\log(x_1) - \log(x_0)]$$

*Note:* This approximation is only mathematically reliable for small differences (typically less than 10%). For larger gaps, the exact formula must be used.
:::

## C4 \[JTRAIN2\]

```{r}
#| label: c4_data

c4 <- jtrain2
kable(head(c4))
```

> (i)

```{r}
#| label: c4_i

c4 |>
  summarise(
    Trained = mean(train)
  ) |>
  round(2)
```

**Interpretation:**

**42%** of men in the sample received training.

> (ii)

```{r}
#| label: c4_ii

c4 |>
  summarise(
    Count = n(),
    Earnings = mean(re78) |> round(2),
    .by = train
  ) |>
  mutate(Train = ifelse(train == 1, "Trained", "Untrained")) |>
  select(Train, Count, Earnings)
```

**Interpretation:**

Trained men received earnings of **6,350** while untrained received **4,550** only. The percentage difference is $\%\Delta x =\dfrac{6.35 - 4.55}{4.55} =  39.56\%$ which is practically significant. A difference of 1800 is economically significant, especially for low-income men in 1978.

::: callout-caution
### Terminology: Change vs. Difference

**Change = Time (Longitudinal)** Tracking the *same* subject across different time periods. \* *Math:* (New - Old) / Old

**Difference = Distinct Groups (Cross-Sectional)** Comparing *separate* subjects at the *same* point in time. \* *Math:* (Group A - Group B) / Group B
:::

> (iii)

```{r}
#| label: c4_iii

c4 |>
  summarise(
    Unemployed = mean(unem78) |> round(2),
    .by = train
  ) |>
  mutate(Train = ifelse(train == 1, "Trained", "Untrained")) |>
  select(Train, Unemployed)
```

**Interpretation:**

**24%** of the trained men were unemployed while **35%** of the untrained men were unemployed. There is a **11 percentage point difference** in unemployment between the two groups.

> (iv)

**Interpretation:**

Training program appears to be effective to increase earnings from part (ii) and decrease unemployment from part (iii). Conclusions will be more convincing if the random sample was balanced, and we controlled for other factors such as earnings and employment before taking the program.

## C5 \[FERTIL2\]

```{r}
#| label: c5_data

c5 <- fertil2
kable(head(c5))
```

> (i)

```{r}
#| label: c5_i

c5 |>
  summarise(
    Min = min(children),
    Max = max(children),
    Average = mean(children)
  ) |>
  round(2)
```

```{r}
#| label: c5_i_2

ggplot(c5, aes(x = factor(children))) +
  geom_bar(fill = "#2c3e50") +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5, size = 3, color = "black") +
  theme_minimal() +
  labs(
    title = "Distribution of Number of Children",
    subtitle = "Based on the c5 dataset observations",
    x = "Number of Children",
    y = "Frequency (Count)",
    caption = "Source: Wooldridge Dataset"
  )
```

**Interpretation**:

Number of children is heavily right skewed. **1132** women had 0 children while only one woman had a maximum number of **13** children.

> (ii)

```{r}
#| label: c5_ii

c5 |>
  summarise(
    Has_Electricity = mean(electric, na.rm = TRUE)
  ) |>
  round(2)
```

**Interpretation**:

**14%** of the women in the sample had home electricity.

> (iii)

```{r}
#| label: c5_iii

c5 |>
  drop_na(electric) |>
  summarise(
    Avg_Children = mean(children) |> round(2),
    .by = electric
  ) |>
  mutate(Home_Electricity = ifelse(electric == 1, "Has Electricity", "No Electricity")) |>
  select(Home_Electricity, Avg_Children)
```

**Interpretation:**

Women who had access to electricity have lower average children **(1.9)** compared to women with no electricity **(2.33)**

> (iv)

**Interpretation:**

No, we cannot conclude that having electricity causes women to have fewer children based solely on this data. While there is a clear negative correlation, this simple comparison in averages suffers heavily from omitted variable bias. Electricity in the home is likely a proxy variable for broader confounding factors—such as higher household income, urban residency, or better access to education and healthcare—which are the more probable underlying drivers of lower fertility rates.

## C6 \[COUNTYMURDERS\]

```{r}
#|label: c6_data

c6 <- countymurders |>
  filter(year == 1996)

kable(head(c6))
```

> (i)

```{r}
#| label: c6_i

c6 |>
  summarise(
    Counties = n_distinct(countyid),
    Zero_Murders = sum(murders == 0),
    Pct_Zero_Executions = mean(execs == 0) * 100
  ) |>
  round(2)
```

**Interpretation:**

There are **2197** counties in the sample. Out of these, **1051** have zero murders, and **98.59%** of them have zero executions.

> (ii)

```{r}
#| label: c6_ii

c6 |>
  summarise(
    Max_Murders = max(murders),
    Max_Execs = max(execs),
    Avg_Execs = mean(execs)
  ) |>
  round(2)
```

```{r}
#| label: c6_ii_2

c6 |>
  tabyl(execs) |>
  adorn_pct_formatting()
```

**Interpretation:**

Largest number of murders is **1403** while largest number of executions is only **3**. Average execution is **0.02** because 98.6% of the sample have zero executions, 1.3% have 1 execution, leaving only 0.1% for 2 executions.

> (iii)

```{r}
#| label: c6_iii

c6 |>
  summarise(
    Correlation = cor(murders, execs)
  ) |>
  round(2)
```

**Interpretation:**

There is a weak correlation **(0.21)** between murders and executions.

> (iv)

**Interpretation:**

No, more executions do not "cause" more murders. The positive, albeit weak, correlation (0.21) is highly misleading for several reasons:

-   **Omitted Variable Bias:** The raw counts ignore population size. Highly populated counties will naturally have higher absolute numbers of both murders and executions compared to rural counties.
-   **Reverse Causality:** The causal arrow likely points the other direction. Counties experiencing higher murder rates might face intense public and political pressure to utilize the death penalty more frequently in response to the crime wave.
-   **Justice System Lag & Attrition:** Executions in 1996 are punishments for crimes committed years or decades prior. Furthermore, many murders go unsolved, meaning the immediate murder count does not directly map to execution rates.
-   **Extreme Data Skew (Zero-Inflation):** Because 98.6% of the counties had exactly zero executions, this data is heavily right-skewed and zero-inflated. A standard linear correlation metric is mathematically inadequate for capturing the true dynamic here.

## C7 \[ALCOHOL\]

```{r}
#| label: c7_data

c7 <- alcohol
kable(head(c7))
```

> (i)

```{r}
#| label: c7_i

c7 |>
  summarise(
    Pct_Alcohol_Abuse = mean(abuse) * 100,
    Employment_Rate = mean(employ)
  ) |>
  round(3)
```

**Interpretation:**

**9.917%** of men in the sample reported abusing alcohol. Self reported employment rate is **0.898** so 89.8% of the men are employed.

> (ii)

```{r}
#| label: c7_ii

c7 |>
  filter(abuse == 1) |>
  summarise(
    Count = n(),
    Employment_Rate = mean(employ)
  ) |>
  round(3)
```

**Interpretation:**

Employment rate for alcohol abusers is **0.873**.

> (iii)

```{r}
#| label: c7_iii

c7 |>
  filter(abuse == 0) |>
  summarise(
    Count = n(),
    Employment_Rate = mean(employ)
  ) |>
  round(3)
```

**Interpretation:**

For non alcohol abusers, employment rate is 0.901.

> (iv)

**Interpretation:**

No, we cannot conclude that alcohol abuse causes unemployment. There is a 3 percentage point difference in the employment rates (87% for abusers versus 90% for non-abusers). However, this is merely a correlation. This simple comparison fails to account for omitted variable bias. Other unobserved factors may be education level, underlying mental or physical health issues.

::: callout-note
### Random Sampling vs. Over-Sampling

A true **random sample** naturally reflects the base rates of the actual population. If only 10% of the real world abuses alcohol, a perfectly random sample `must` show a roughly 9-to-1 imbalance.

**Over-sampling** happens when researchers intentionally skew the collection process to gather more data on a minority group (e.g., artificially forcing a 50/50 split).

Therefore, massively unequal group sizes in your data are a sign of healthy randomization.
:::

## C8 \[ECONMATH\]

```{r}
#| label: c8_data

c8 <- econmath
kable(head(c8))
```

> (i)

```{r}
#| label: c8_i

c8 |>
  summarise(
    Count = n(),
    HS_Economics = sum(econhs)
  )
```

**Interpretation:** The sample has **856** students. Among those students, **317** took the economics class in high school.

> (ii)

```{r}
#| label: c8_ii
c8 |>
  summarise(
    Avg_Score = round(mean(score), 2),
    .by = econhs
  ) |>
  mutate(HS_Economics = ifelse(econhs == 1, "HS Economics", "No HS Economics ")) |>
  select(HS_Economics, Avg_Score)
```

**Interpretation:** Average score for students with no high school economics is **72.91** while students with high school economics scored on average **72.08**

> (iii)

No, students with high school economics chose to take the course in high school so we have selection bias. There is a small difference between the two groups that may be due to other factors or pure chance.

> (iv)

**Interpretation:** I would conduct an **RCT** in which I randomly assign students to high school economics and other students stay as control. Then I will track both groups score in the university course.

## C9 \[LOANAPP\]

```{r}
#| label: c9_data

c9 <- loanapp
kable(head(c9))
```

> (i)

```{r}
#| label: c9_i

c9 |>
  summarise(
    Pct_White = mean(white) * 100
  ) |>
  round(2)
```

**Interpretation:** **84.51%** of the applicants in the sample are white.

> (ii)

```{r}
#| label: c9_ii

c9 |>
  filter(white == 0) |>
  summarise(
    Approved_Pct = mean(approve) * 100
  ) |>
  round(2)
```

**Interpretation:**

Among non-white applicants, **70.78%** have their application approved.

> (iii)

```{r}
#| label: c9_iii

c9 |>
  filter(white == 1) |>
  summarise(
    Approved_Pct = mean(approve) * 100
  ) |>
  round(2)
```

**Interpretation:**

Among white applicants, **90.84%** have their application approved.

> (iv)

**Interpretation:**

The percentage point difference is $90.84 - 70.78 = 20.06$. Although the gap between both groups is big, it can be due to other factors like employment status and crime record.

::: callout-warning
### Percentage Difference vs. Percentage Point Difference

**Percentage Point Difference:** The absolute mathematical difference between two rates.

$$\text{Difference in Points} = \text{Rate}_A - \text{Rate}_B$$

**Percentage Difference:** The relative size of the gap compared to a chosen base group.

$$\% \text{Difference} = \frac{\text{Rate}_A - \text{Rate}_B}{\text{Rate}_B} \times 100$$
:::
