---
title: "Multiple Regression Analysis: Estimation"
format: html
---

## Computer Exercises

```{r}
#| label: 'setup'
#| include: false
library(tidyverse)
library(wooldridge)
library(modelsummary)
```

### C1 \[BWGHT\]

```{r}
#| label: c1_data
df_bwght <- wooldridge::bwght
knitr::kable(head(df_bwght))
```

> (i)

**Interpretation:**

$\beta_2$ is expected to be **positive**. Higher family income $\rightarrow$ better nutrition $\rightarrow$ higher birth weight

> (ii)

**Interpretation:**

Yes, **negative** correlation. smoking rates is higher among lower-income demographics. Omitting family income will cause Omitted Variable Bias (OVB).

> (iii)

```{r}
#| label: c1_iii
model_bwght_simple <- lm(bwght ~ cigs, data = df_bwght)
model_bwght_multiple <- lm(bwght ~ cigs + faminc, data = df_bwght)

models_bwght <- list(
  "Without Income" = model_bwght_simple,
  "With Income" = model_bwght_multiple
)

modelsummary(
  models_bwght,
  fmt = 2,
  statistic = NULL,
  gof_map = c("nobs", "r.squared"),
  coef_rename = c(
    "cigs" = "Cigarettes/Day", 
    "faminc" = "Family Income ($1000s)", 
    "(Intercept)" = "Constant"
  )
)
  
```

**Interpretation:**

Including income reduces the smoking coefficient from approximately $-0.514$ to $-0.463$. This indicates that the simple regression suffered from omitted variable bias. 
$$\text{Bias} = \beta_2 \times \tilde \delta_1$$

1\. $\beta_2$: Positive, rich people have heavier babies.

2\. $\tilde \delta_1$: Negative, poor people smoke more

3\. Effect: Positive $\times$ Negative = **Negative Bias**

4\. $\beta_1$ is negative, negative bias will increase it from -0.46 to -0.51

### C2 \[HPRICE1\]

```{r}
#| label: c2_data
df_hprice1 <- wooldridge::hprice1
knitr::kable(head(df_hprice1))
```

> (i)

```{r}
#| label: c2_i
model_hprice <- lm(price ~ sqrft + bdrms, data = df_hprice1)

modelsummary(
  model_hprice,
  fmt = 2,
  statistic = NULL,
  gof_map = c("nobs", "r.squared"),
  coef_rename = c(
    "(Intercept)" = "Constant",
    "sqrft"       = "Size (sq. ft.)",
    "bdrms"       = "Number of Bedrooms"
  )
)
```

> (ii)

**Interpretation:**

Holding square footage constant, adding one more bedroom increases price by \$15,200. This estimate is not useful because adding small rooms without adding square footage will make it feel cramped.

> (iii)

```{r}
#| label: c2_iii
base_house <- data.frame(sqrft  = 140, bdrms = 1)
upgraded_house <- data.frame(sqrft  = 280, bdrms = 2)

price_1 = predict(model_hprice, newdata = base_house)
price_2 = predict(model_hprice, newdata = upgraded_house)

price_increase <- price_2 - price_1

paste(round(price_increase, 2))
```

**Interpretation:**

The estimates increase is **\$33,120**. This represents the combined value of adding a room that is 140 square feet 
$$140 \times 0.13 + 15.2 = 33.2$$

> (iv)

**Interpretation:**

**63.2%** of the variation in house prices are explained by the two variables.

> (v)

```{r}
#| label: c2_v
new_data <- data.frame(sqrft = 2438, bdrms = 4)
price = predict(model_hprice, newdata = new_data)
paste(round(price, 2))
```

or

```{r}
#| label: c2_v2
result <- model_hprice$fitted.values[1]
paste(round(result, 2))
```

or

```{r}
#| label: c2_v3
price = coef(model_hprice)["(Intercept)"] +
  (coef(model_hprice)["sqrft"] * 2438) + 
  (coef(model_hprice)["bdrms"] * 4)
paste(round(price, 2))

```

**Interpretation:**

Predicted selling price is **\$354,000**

> (vi)

```{r}
#| label: cs_vi
residual <- model_hprice$residuals[1]
paste(round(residual, 2))
```

**Interpretation:**  Residual is **\$-54,610**. The buyer underpaid for the house.

### C3 \[CEOSAL2\]

```{r}
#| label: c3_data
df_ceosal2 <- wooldridge::ceosal2
knitr::kable(head(df_ceosal2))
```

> (i)

```{r}
#| label: c3_i
model_ceosal2 <- lm(lsalary ~ lsales + lmktval, data = df_ceosal2)

modelsummary(
  model_ceosal2,
  fmt = 2,
  statistic = NULL,
  gof_map = c("nobs", "r.squared"),
  coef_rename = c(
    "(Intercept)" = "Constant",
    "lsales"  = "Log(Sales)",
    "lmktval" = "Log(Market Value)"
  )
)
```

> (ii)

```{r}
#| label: c3_ii
ggplot(df_ceosal2, aes(x = profits)) +
  geom_histogram(bins = 30)
```

```{r}
#| label: c3_ii2
model_ceosal2_2 <- lm(lsalary ~ lsales + lmktval + profits, data = df_ceosal2)

modelsummary(
  model_ceosal2_2,
  fmt = 2,
  statistic = NULL,
  gof_map = c("nobs", "r.squared"),
  coef_rename = c(
    "(Intercept)" = "Constant",
    "lsales"  = "Log(Sales)",
    "lmktval" = "Log(Market Value)",
    "profits" = "Profits"
  )
)
```

**Interpretation:**

Profits can't be added in log form because it has non positive values. The model explains **30%** of the variation in the salaries. The remaining 70% is determined by other factors like tenure, unobserved CEO talent, or corporate politics.

> (iii)

```{r}
#| label: c3_iii
model_ceosal2_2 <- lm(
  lsalary ~ lsales + lmktval + profits + ceoten, data = df_ceosal2)

modelsummary(
  model_ceosal2_2,
  fmt = 2,
  statistic = NULL,
  gof_map = c("nobs", "r.squared"),
  coef_rename = c(
    "(Intercept)" = "Constant",
    "lsales"  = "Log(Sales)",
    "lmktval" = "Log(Market Value)",
    "profits" = "Profits",
    "ceoten" = "Tenure"
  )
)
```

**Interpretation:**

$\beta_4$ = 0.01, one additional year of tenure is associated with **1%** increase in salary, holding firm performance constant.

> (iv)

```{r}
#| label: c3_iv
df_ceosal2 |>
  summarise(
    Correlation = cor(lmktval, profits)
  ) |>
  round(2)
```

**Interpretation:**

Correlation = **0.78**. This high correlation between market value and profits indicates multicollinearity. While this does not bias our estimates or inflate the overall $R^2$, it makes it difficult to isolate the individual partial effects of profits versus market value.
