---
title: "Multiple Regression Analysis: OLS Asymptotics"
format: html
---


```{r}
#| label: 'setup'
#| message: false
#| warning: false

library(tidyverse)
library(wooldridge)
library(knitr)
library(modelsummary)
library(broom)
library(car)
```

## C1 [WAGE1]

> Use the data in WAGE1 for this exercise.

```{r}
#| label: c1_data

df_wage1 <- wooldridge::wage1
knitr::kable(head(df_wage1))
```

> (i) Estimate the equation and plot the residuals $$wage = \beta_0 + \beta_1 educ + \beta_2 exper + \beta_3 tenure + u$$ Save the residuals and plot a histogram.

```{r}
#| label: c1_i

model_wage1 <- lm(
  wage ~ educ + exper + tenure,
  data = df_wage1
)

ggplot(data = model_wage1, aes(x = .resid)) +
  geom_histogram(bins = 30)
```

**Interpretation:**

The residuals are highly right-skewed and clearly violate the assumption of normality.

> (ii) Repeat part (i), but with log(`wage`) as the dependent variable.

```{r}
#| label: c1_ii

model_lwage <- lm(
  lwage ~ educ + exper + tenure,
  data = df_wage1
)

ggplot(data = model_lwage, aes(x = .resid)) +
  geom_histogram(bins = 30)
```

**Interpretation:**

Residuals of the log wage model seem symmetric with few outliers.

> (iii) Would you say that Assumption MLR.6 is closer to being satisfied for the level-level model or the log-level model?

```{r}
#| label: c1_iii

shapiro_level <- shapiro.test(resid(model_wage1))
print(shapiro_level)

shapiro_log <- shapiro.test(resid(model_lwage))
print(shapiro_log)
```

**Interpretation:**

The log-level model is much closer to satisfying MLR.6. The level-level residuals are heavily right-skewed, blatantly violating the normality assumption. By taking the log of wages, the distribution of residuals becomes roughly symmetric and bell-shaped, which aligns much more closely with the normality requirement of MLR.6.

::: callout-tip
using `model$residuals` will not work with all models. Instead use `resid(model)`

To get coefficients use `tidy(model)`

To get Model-Level Statistics use `glance(model)`

To get Observation-Level Statistics use `augment(model)`

For presentation, keep using `modelsummary(model)`
:::

## C2 [GPA2]

> Use the data in GPA2 for this exercise

```{r}
#| label: c2_data

df_gpa2 <- wooldridge::gpa2
knitr::kable(head(df_gpa2))
```

> (i) Using all 4,137 observations, estimate the equation $$colgpa = \beta_0 + \beta_1 hsperc + \beta_2 sat +u$$ and report the results in standard form.

```{r}
#| label: c2_i

model_gpa2 <- lm(
  colgpa ~ hsperc + sat,
  data = df_gpa2
)

modelsummary(
  model_gpa2,
  fmt = 2,
  shape = term ~ model + statistic,
  estimate = "{estimate}{stars}",
  statistic = c("std.error", "statistic", "p.value"),
  gof_map = c("nobs", "r.squared", "adj.r.squared"),
  coef_rename = c(
    "(Intercept)" = "Constant",
    "hsperc" = "High School Percentage",
    "sat" = "Sat Score"
  )
)
```

> (ii) Reestimate the equation using first 2070 observations.

```{r}
#| label: c2_ii

df_gpa2_sub <- df_gpa2 |>
  slice(1:2070)

model_gpa2_sub <- lm(
  colgpa ~ hsperc + sat,
  data = df_gpa2_sub
)

modelsummary(
  model_gpa2_sub,
  fmt = 2,
  shape = term ~ model + statistic,
  estimate = "{estimate}{stars}",
  statistic = c("std.error", "statistic", "p.value"),
  gof_map = c("nobs", "r.squared", "adj.r.squared"),
  coef_rename = c(
    "(Intercept)" = "Constant",
    "hsperc" = "High School Percentage",
    "sat" = "Sat Score"
  )
)
```

> (iii) Find the ratio of the standard errors on `hsperc` from parts (i) and (ii).

```{r}
#| label: c2_iii

se_full <- tidy(model_gpa2) |>
  filter(term == "hsperc") |>
  pull(std.error)

se_sub <- tidy(model_gpa2_sub) |>
  filter(term == "hsperc") |>
  pull(std.error)

print(se_sub / se_full)
```

**Interpretation:**

The ratio is approximately 1.3. This is close to the theoretical penalty of $\sqrt{2} \approx 1.414$. The penalty is not exactly 1.414 because while subsetting the data cut $SST_j$ roughly in half, it also caused slight fluctuations in the residual variance ($\hat{\sigma}$) and the variable's collinearity ($R^2_j$). $$SE(\hat \beta_j) = \frac{\hat \sigma}{\sqrt{SST_j(1-R^2_j)}}$$

## C3 [BWGHT]

> In equation (4.42) of Chapter 4, using the data set BWGHT, compute the LM statistic for testing whether `motheduc` and `fatheduc` are jointly significant. In obtaining the residuals for the restricted model, be sure that the restricted model is estimated using only those observations for which all variables in the unrestricted model are available (see Example 4.9).

```{r}
#| label: c3_data

df_bwght <- wooldridge::bwght
knitr::kable(head(df_bwght))
```

> compute the LM statistic for testing whether motheduc and fatheduc are jointly significant. $$\begin{align}bwght =& \beta_0 + \beta_1 cigs + \beta_2 parity + \beta_3 faminc \\&+ \beta_4 motheduc + \beta_5 fatheduc + u\end{align}$$

**The Math Way:**

To do Lagrange Multiplier Test:

1.  Filter the data from missing data
2.  Estimate the restricted model
3.  Extract the residuals $\tilde u$
4.  Regress $\tilde u$ on all independent variables
5.  Extract $R^2_u$ and $n$
6.  $LM = n \cdot R^2_u$
7.  $LM \sim \chi^2_q$ where q is number of restrictions

```{r}
#| label: c3_i

df_clean <- df_bwght |>
  drop_na(bwght, cigs, parity, faminc, motheduc, fatheduc)

model_r <- lm(bwght ~ cigs + parity + faminc,
  data = df_clean
)

df_clean$u_tilde <- resid(model_r)

model_aux <- lm(
  u_tilde ~ cigs + parity + faminc + motheduc + fatheduc,
  data = df_clean
)

n <- nobs(model_aux)
r2_aux <- summary(model_aux)$r.squared

lm_stat <- n * r2_aux
p_val <- pchisq(lm_stat, df = 2, lower.tail = FALSE)

cat("LM Statistic: ", lm_stat, "\n")
cat("P-value:      ", p_val, "\n")
```

**Interpretation:**

Calculated LM Statistic is 2.8 with p value 0.24. We fail to reject the null hypothesis; education is not jointly statistically significant.

::: callout-important
If we wanted to test if education = 0.5 not just zero, we will modify y from step two

`df_clean <- df_clean |>  mutate(bwght_tilde = bwght - (0.5 * motheduc) - (0.5 * fatheduc))`
:::

## C4 [SKEWNESS]

> Several statistics are commonly used to detect nonnormality in underlying population distributions. Here we will study one that measures the amount of skewness in a distribution. Recall that any normally distributed random variable is symmetric about its mean; therefore, if we standardize a symmetrically distributed random variable, say $z = (y - \mu_y)/\sigma_y$, where $\mu_y = \text{E}(y)$ and $\sigma_y = \text{sd}(y)$, then $z$ has mean zero, variance one, and $\text{E}(z^3) = 0$. Given a sample of data $\{y_i: i = 1, \dots, n\}$, we can standardize $y_i$ in the sample by using $z_i = (y_i - \hat{\mu}_y)/\hat{\sigma}_y$, where $\hat{\mu}_y$ is the sample mean and $\hat{\sigma}_y$ is the sample standard deviation. (We ignore the fact that these are estimates based on the sample.) A sample statistic that measures skewness is $n^{-1} \sum_{i=1}^n z_i^3$, or where $n$ is replaced with $(n - 1)$ as a degrees-of-freedom adjustment. If $y$ has a normal distribution in the population, the skewness measure in the sample for the standardized values should not differ significantly from zero.

> (i) First use the data set 401KSUBS, keeping only observations with `fsize = 1`. Find the skewness measure for `inc`. Do the same for `log(inc)`. Which variable has more skewness and therefore seems less likely to be normally distributed?

> (ii) Next use BWGHT2. Find the skewness measures for `bwght` and `log(bwght)`. What do you conclude?

> (iii) Evaluate the following statement: “The logarithmic transformation always makes a positive variable look more nearly normally distributed.”

> (iv) If we are interested in the normality assumption in the context of regression, should we be evaluating the unconditional distributions of $y$ and $\log(y)$? Explain.

## C5 [HTV]

> Consider the analysis in Computer Exercise C11 in Chapter 4 using the data in HTV, where `educ` is the dependent variable in a regression.

> (i) How many different values are taken on by `educ` in the sample? Does `educ` have a continuous distribution?

> (ii) Plot a histogram of `educ` with a normal distribution overlay. Does the distribution of `educ` appear anything close to normal?

> (iii) Which of the CLM assumptions seems clearly violated in the model
> $$educ = \beta_0 + \beta_1 motheduc + \beta_2 fatheduc + \beta_3 abil + \beta_4 abil^2 + u$$?
> How does this violation change the statistical inference procedures carried out in Computer Exercise C11 in Chapter 4?

## C6 [ECONMATH]

> Use the data in ECONMATH to answer this question.
> (i) Logically, what are the smallest and largest values that can be taken on by the variable `score`? What are the smallest and largest values in the sample?
> (ii) Consider the linear model
> $$score = \beta_0 + \beta_1 colgpa + \beta_2 actmth + \beta_3 acteng + u.$$
> Why can Assumption MLR.6 not hold for the error term $u$? What consequences does this have for using the usual $t$ statistic to test H$_0$: $\beta_3 = 0$?
> (iii) Estimate the model from part (ii) and obtain the $t$ statistic and associated $p$-value for testing H$_0$: $\beta_3 = 0$. How would you defend your findings to someone who makes the following statement: “You cannot trust that $p$-value because clearly the error term in the equation cannot have a normal distribution.”

## C7 [APPLE]

> Use the data in APPLE to answer the following questions.

> (i) Estimate the following equation by OLS, and report the results in the usual format, including standard errors, sample size, and $R$-squared: $$ecolbs = \beta_0 + \beta_1 ecoprc + \beta_2 regprc + \beta_3 faminc + \beta_4 age + \beta_5 educ + u$$ Are any of the coefficients on `faminc`, `age`, and `educ` individually statistically significant? Do any seem practically large?

> (ii) Obtain the $F$ statistic for joint significance of `faminc`, `age`, and `educ`; report its value, and the associated $p$-value, to three decimal places. What do you conclude?

> (iii) Estimate the equation without the variables `faminc`, `age`, and `educ`. Obtain the residuals, $\tilde{u}_i$. If you regress these residuals on $ecoprc_i$ and $regprc_i$, what is the $R$-squared, and why?

> (iv) Now use the residuals from part (iii) to obtain the $LM$ statistic for joint significance of `faminc`, `age`, and `educ`. Report the statistic to three decimal places, and likewise with the associated $p$-value. How similar is your conclusion to when you use the $F$ statistic?
