# Preface {.unnumbered}



This digital book serves as a comprehensive computational companion to Jeffrey Wooldridge’s Introductory Econometrics: A Modern Approach. As a Teaching Assistant at Alexandria University, I developed this resource to bridge the gap between theoretical econometric concepts and practical data analysis using the R programming language.

This project features detailed solutions to 279 end-of-chapter exercises, providing a rigorous application of econometric theory to real-world datasets. The scope of this work covers a diverse range of modeling frameworks, including:

Linear Regression: From Simple OLS to Multiple Regression Analysis.

Qualitative Variables: Binary (Dummy) variables and Linear Probability Models (LPM).

Limited Dependent Variables: Logit, Probit, and Tobit models.

Time Series Analysis: Static models, Finite DLMs, and Serial Correlation.

Panel Data Methods: Pooled OLS, Fixed Effects, and Random Effects.

Causal Inference: Instrumental Variables (IV), Two-Stage Least Squares (2SLS), and Difference-in-Differences (DiD).

## About the Author
I am Ahmed H. Darwish, a Teaching Assistant with a Bachelor’s degree in Business Administration, majoring in Applied Statistics. Currently, I am pursuing my Master’s degree and conducting research in the field of econometrics.

My work focuses on:

Applied Econometrics: Replicating complex academic models to ensure reproducibility.

Causal Inference: Applying modern frameworks from The Effect to real-world data.

Open Source Contribution: I am a contributor to the `wooldridge` Python library.

## Exercises Solved

```{r}
#| label: progress
#| message: false
#| warning: false
#| code-fold: true
#| code-summary: "Click to see how the tracker is generated"

library(tidyverse)
library(gt)
library(gtExtras)

progress_data <- tribble(
  ~Chapter, ~Questions, ~Solved,
  1, 9, 9,
  2, 12, 12,
  3, 14, 7,
  4, 15, 3,
  5, 7, 3,
  6, 16, 2,
  7, 19, 0,
  8, 14, 0,
  9, 15, 0,
  10, 15, 0,
  11, 15, 0,
  12, 17, 0,
  13, 18, 0,
  14, 20, 0,
  15, 15, 0,
  16, 12, 0,
  17, 18, 0,
  18, 18, 0,
  19, 10, 0
)

stats <- progress_data |>
  summarise(
    Total = sum(Questions),
    Solved = sum(Solved),
    Percent = (Solved / Total)
  )


ggplot(stats) +
  geom_rect(aes(xmin = 0, xmax = Total, ymin = 0, ymax = 1), fill = "grey90") +
  geom_rect(aes(xmin = 0, xmax = Solved, ymin = 0, ymax = 1), fill = "#4CAF50") +
  geom_text(
    aes(
      x = Total / 2, y = 0.5,
      label = paste0(round(Percent * 100, 1), "% Complete (", Solved, "/", Total, ")")
    ),
    fontface = "bold", size = 6
  ) +
  theme_void() +
  labs(title = "Overall Book Progress") +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18, margin = margin(b = 10)),
    plot.margin = margin(20, 20, 20, 20)
  )
```


```{r}
#| label: progress
#| message: false
#| warning: false
#| code-fold: true
#| code-summary: "Click to see how the tracker is generated"


progress_data <- progress_data |>
  mutate(
    Percentage = Solved / Questions,
    Status = case_when(
      Percentage == 1 ~ "Finished",
      Percentage > 0 ~ "In Progress",
      TRUE ~ "Pending"
    )
  )


progress_data |>
  gt() |>
  gt_plt_bar_pct(column = Percentage, scaled = FALSE, fill = "#2c3e50", background = "#ecf0f1") |>
  gt_badge(Status, palette = c("Finished" = "#27ae60", "In Progress" = "#f39c12", "Pending" = "#bdc3c7")) |>
  cols_label(
    Percentage = "Progress",
    Solved = "Completed"
  ) |>
  tab_header(
    title = md("**Exercise Completion Tracker**"),
    subtitle = "Real-time status of the Wooldridge Solution Manual"
  ) |>
  tab_options(
    table.width = pct(90),
    column_labels.font.weight = "bold",
    heading.title.font.size = px(24)
  ) |>
  cols_align(align = "center", columns = everything())
```


::: callout-note
8th edition increased computer exercises from 246 to 279
:::

------------------------------------------------------------------------

*This project is a work in progress as part of my Master's thesis portfolio. For collaborations or inquiries regarding remote data analysis, feel free to contact me via GitHub.*
